### 2022-05-10 paper/manuscript.Rmd
trilemma manuscript v May 2022

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
# author: |
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
abstract: |
  Sustainability is often framed as achieving prosperity, equality, and low environmental impact. But what if there is a trilemma that inhibits the joint achievement of these three key goals? Here we analyzed the synergies and tradeoffs between these goals by investigating historical data on per-capita gross national income (prosperity), the Gini coefficient for income distribution (inequality), and per-capita ecological footprint (environmental impact). We find that no country in the world have successfully solved the trilemma, but there are common typologies that can inform common barriers and strategies to solve it. Some countries are trapped on a high inequality space, while prosperous countries can reduce inequality but typically at expenses of high environmental impacts. Achieving ambitous sustainable goals will require redistribution of wealth and responsibility for the stewardship of the planet.

output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
csl: /Users/juanrocha/Documents/styles/nature.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
---

```{r data, include=FALSE}
library(tidyverse)
library(here)

```

Targets:

- NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
70max

- WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
abstract

Current: 3957w

# Introduction

(Aim 1000w | Current 671w)

Sustainability has often been characterised as having three essential, nested dimensions: economy, society, and the environment. More specifically, sustainability has been defined as attaining high material living standards (prosperity), that are also well distributed (equality) and maintainable over the long run (achieved through environmental integrity”) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Although each is undeniably desirable for development, what if all three goals are not simultaneously achievable? It has often been assumed that the ends to which individuals as well as societies aspire inevitably cohere, but *a priori*, there is no reason to believe that diverse goals must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
@Greenspan:2001].

In the literature on development, such macro-level incompatibilities are not novel. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
[@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011]. Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in improving diet, mitigating diseases, and environmental integrity [@Clark:2018]. Another such “trilemma” that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.

For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
@Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, approximately 10% of the world’s population still lives under the international poverty line (less than US\$ 1.90 dollars per day), with a much larger share subsisting above that threshold but still aspiring for higher and more secure living standards
[@TheWor:2018]. In a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.

At the same time, while many average incomes for countries in what was once called the “Third World” have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that such unequal distribution arouses has historically been called the "Social Question"[@Faist:2009]. Over the past decade, inequality has become an ever-urgent issue for policymakers as well as the general public [@Chancel:2020]. This urgency has been strengthened by a growing body of research demonstrating links between inequality and human wellbeing, health, environmental quality, and economic complexity [@Hamann:2018; @Leach:2018;
@Pikett:2014; @Wilkinson:2009; @Hartmann:2017].

Environmental conditions form the third, and arguably most fundamental, dimension of the sustainability trilemma. There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere [@Díaz:2019]. Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
[@Díaz:2019]. At the same time, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017]. Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].

Although all three goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two. Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly prioritize prosperity, equality, and the environment [@Sachs:2015]. In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by a given country and by the world as a whole. In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, and if it does exist, where does it occur and why does it occur there?

# Methods

(Aim 800w | current 972w)

To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above  (Table \ref{tab1}). We chose the country scale due to practical considerations
of data availability, but also on the recognition that many of the most
consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018]. We operationalized the trilemma as the movement of countries overtime in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g. reducing inequality and increasing prosperity) There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies are movements along
the (blue) diagonal where the achievement of two goals can be positively correlated;
while tradeoffs are represented by the (red) axis where the achievement of one
pervades the achievement of another (negative correlations). It is important to note that in this framing, the sign of the correlation changes depending on the term used: i.e., if one speaks of equity as opposed to inequality, or environmental quality as opposed to environmental footprint. In the remainder of this paper we use inequality and environmental footprint because that is how the original data was structured. Therefore, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 4in]{conceptual_fig}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality while increasing prosperity and environmental quality. There could be synergies and trade-offs in achieving these three objectives. We explore trajectory typologies on the trilemma space.}
\label{fig:trilemma}
\end{figure*}
```

To quantify prosperity, equality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used
per-capita gross national income (GNI) in US\$ values.GNI is a cognate measure of gross domestic product (GDP) and the two are close in value for most countries. A key difference is that GNI accounts for earnings from foreign investments. The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source:
<http://data.worldbank.org>). The threshold for being "prosperous" was   US\$12,746 per capita per year, which is the World Bank’s definition for a high-income country [@TheWor:2018].

To measure inequality, we used the Gini coefficient for income. The Gini coefficient is a statistical measure of dispersion and has a long history of being used to measure the level of economic inequality within a country [@Pyatt:1976]. Hypothetically, a country with a Gini value of 0 is absolutely equal, while a value of 1 represents absolute inequality. Gini coefficients of income were taken from the World Income Inequality Database v. 3.4 (source: [https://www.wider.unu.edu/data](https://www.wider.unu.edu/data)). We used 0.30 as the threshold below which countries are identified as having low inequality, similar to findings from a historical survey by Piketty [@Piketty:2014].

We used per-capita ecological footprint to measure environmental impact. Ecological footprint is a holistic estimate of the biophysical resources used by human activity, in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998]. Its unit is the "global hectare", combining measurements of carbon emissions, built-up area, fishing grounds, croplands, forests, and grazing land into an index of environmental impact. In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological “deficit”. Though it is not exhaustive, the ecological footprint provides the most widely-used and data-rich approximation of overall human impact on the environment, and it accounts for the effects of trade on resource consumption by country. Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)). We used 1.68 global per capita hectares as a threshold for environmental impact, as it reflects the amount of average land needed to satisfy human needs without exceeding the global biocapacity to recover. In other words, it is the limit at which current generations’ needs can be satisfied without compromising the ability of future generations to do so. Ecological footprint estimates above 1.68 global per capita hectares, therefore, represent unsustainable levels of environmental impact.

```{r tab1, echo = FALSE, include = TRUE, results = "asis"}
dat <- tibble(
    Dimension = c("Prosperity", "Equality", "Environment"),
    Indicator = c(
        "Per-capita gross national income in US$",
        "Gini coefficient for income",
        "Per-capita ecological footprint in global hectares"),
    N = c(232, 196, 211),
    Threshold = c(12746, 0.3, 1.68)
)

stargazer::stargazer(
    dat, type = "latex", 
    title = "Indicators used in the assessment of national prosperity, equality and environmental impact.",
    label = "tab1", summary = FALSE, header = FALSE)

# knitr::kable(dat, caption = "Indicators used in the assessment of national prosperity, equality and environmental impact.", label = "tab1")
```

We used these data to chart the movement of countries across dimensions of prosperity, equality, and environmental integrity over time. We further investigated whether these movements trace out typologies of countries in the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters to fit. We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization purposes, and vector fitting to show in the principal component space how different features of the trilemma space pull clusters apart. All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).

An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces is only possible for 53 countries. The most incomplete dataset is the inequality data (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemmma space (Fig \ref{fig:cube_time}).

# Results

(Aim 1200w | 610w)

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{cube_time}
\caption{\textbf{Countries trajectories in the trilemma space} 53 trajectories from 1995 to 2017 are depicted for countries with $\leqslant 30\%$ missing values. The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (A), inequality and prosperity (B), and ecological footprint and prosperity (C). Thresholds that delimit the objective space are depicted as dotted lines.}
\label{fig:cube_time}
\end{figure*}
```

No country in the data simultaneously achieved high prosperity, low inequality, and high environmental standards. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between inequality and ecological footprint (Fig \ref{fig:cube_time}A - as the Gini indicator for income decreases, ecological footprint increases) and ecological footprint and prosperity (Fig \ref{fig:cube_time}C- ecological footprint moves in the same direction as per-capita GNI). Additionally, prosperity and inequality (Fig \ref{fig:cube_time}B) were also correlated, with a number of high-income countries (e.g., Sweden, New Zealand) having low levels of inequality in the past, but then showed high variation and a trend toward higher inequality levels in more recent years.

While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typology. To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ is the angle in the inequality and prosperity plane while $\theta$ is the angle in the inequality and ecological footprint plane. These two angles and the total distance covered in the three-dimensional space are used to uncover country typologies.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{trilema_ensamble}
\caption{\textbf{Trilemma attractors} }
\label{fig:cube_ensamble}
\end{figure*}
```

There are certain regions of the trilemma space where countries tend to dwell (Fig \ref{fig:cube_ensamble}). The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three clusters of countries.  Despite this clustering, countries showed distinctive profiles in terms of their direction of movement in the trilemma space (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's inequality and ecological footprint worsened while GNI per capita grew significantly. And while in Ukraine inequality and environmental impact actually fell, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}). First, there was a group of countries trapped in a region of high inequality, mainly in Latin America. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$, that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but with no definite direction, characterized by particularly high variations in Gini and per-capita GNI (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group). If we ignore temporal dynamics and focus on the mean and standard deviation as summary statistics, we are able to include more countries (N=140). Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and the environment. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3in]{pca_trilemma}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. 3 clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A). 2 clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). }
\label{fig:pca}
\end{figure*}
```

# Discussion

(Aim 1000 | 1472)

Our analyses found that inequality and ecological footprint goals pull countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time. This suggests the existence of regions within the trilemma space that are hard to escape, characterized by high levels of inequality. The historical data shows that countries typically reduce inequality and then increase ecological footprint, not the other way around (Fig \ref{fig:cube_ensamble}A). Importantly, many middle-income countries have similar levels of inequality as high-income ones, but at lower levels of ecological footprint and per-capita GNI (Fig \ref{fig:cube_ensamble}B, C). This means that countries can achieve low levels of inequality while attaining and preserving high environmental quality. Historical evidence shows that higher prosperity often leads to higher ecological footprints. Nonetheless, it is important to note that past trajectories do not necessarily determine plausible future pathways in the trilemma space.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 1.5in]{maps_trilemma}
\caption{\textbf{Countries mean performance on trilemma space} Mean values of ecological footprint (A), inequality (B), and gross national income (C). }
\label{fig:map}
\end{figure*}
```
The potential mechanisms of the tradeoffs and synergies between economic, social, and environmental goals remain open questions. Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the later grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. Since this relationship was first posited in 1955, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020]. This has made the identification of mechanisms difficult.  Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through systemic regularities of the global economy or proactive policies by national governments [@Pikett:2014]. In our analysis of the trilemma space, we did not find evidence for a KC-like relationship.

In terms of the economy-environment interface, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship and implicit explanation: countries increase their environmental footprints as incomes rise (which in modern history has been driven by industrializatoin and urbanization); but after reaching an inflection point of affluence, the footprints of an increasingly rich society decline (Barbier 1997). As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not (Dasgupta et al. 2002; Stern 2004). In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC. Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns (Mäler 2013). But what is clear from the wealth of diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.

Other mechanisms and pathways that do not posit linear or quadratic relationships have been identified as driving dynamics between economic, social and environmental dimensions. For instance, countries with high per-capita biocapacity tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions [@Weinzett:2013]. Similarly, population, environmental carrying capacity, and pollution exhibit mutually-reinforcing feedbacks across countries [@LaTorr:2019]. These types of feedbacks can create and maintain "red loop" or "green loop" dynamics at the national level, entrapping some states in long-term conditions of environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018; @Hamann:2015].

Additionally, structural forces in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens. Environmental impacts can be “outsourced”, as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses (Mäler 2013). For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013]. Additionally, it has been shown that international trade has significantly influenced global land use patterns, worsening natural resource exploitation and biodiversity loss [@Creutzig:2019]. Thus telecoupling and resource consumption from far away places is a mechanism that can maintain countries in undesirable regions of the trilemma space.

While no country was found to have eluded the sustainability trilemma, there were states, forming clusters, in which tradeoffs and synergies between economic, social and environmental goals moved in distinctively different directions. These dynamics further highlight the inadequacy of reductionistic, often mono-causal, explanations - such as those found in many KC and EKC studies - and the need to incorporate historical, geographic, and structural factors.  For instance, in Ukraine, the existence of (relatively) low inequality and environmental footprint alongside low prosperity seemed to conform to the KC and EKC relationships. However, inequality levels in the Philippines, which had a similar average income, and Mexico, which had a similar per-capita ecological footprint, were extremely high. Additionally, Ukraine’s trajectory through the trilemma space was far from smooth: over the study period, average incomes rose and fell in dramatic downswings due to corruption (endemic and structural across post-Soviet Union countries) and war, among other historical factors.

The clustering of countries in the trilemma space also shows the importance of geographic factors, especially territorial features and natural resource endowments, to sustainable development.  Countries that performed relatively well along all three dimensions tended to be advanced Western economies, particularly along the North Atlantic rim (e.g., Canada, Sweden) and in the Antipodes (i.e., New Zealand and Australia). These countries, in addition to benefiting from the legacies of early industrialization and imperialism, also had a much higher ratio of space and resources to people, as reflected for instance in per-capita biocapacity. Finally, the structure of the global economy is also a powerful factor in explaining international differences in the trilemma space. China, the rapid rise of both prosperity, inequality, and environmental footprint were significantly driven by its integration into international networks of commerce and finance (most notably its accession to the World Trade Organization in 2002). Rapid economic growth was driven by the emergence of a resource-intensive manufacturing sector - with China becoming the so-called “factory of the world”. At the same time, studies show that the surge of economic inequality in China was due in large part to geographic differences: coastal areas were the manufacturing base and export hubs, with high-value added contributions, while the inland areas supplied mostly raw materials, intermediate goods, and cheap labour.

Although we believe our analysis of the sustainability trilemma identified important patterns and drivers of international development (as well as lacunae in the existing literature), it still faced limitations of data and method. First, the offshoring of environmental degradation discussed above is also not completely captured in standard ecological footprint measurements [@VandenBergh:2012].  A bottom-up or consumption-based approach to measuring environmental impacts – as opposed to the more mainstream top-down or production-based approach – may help internalize these externalities (e.g., resource-intensive goods produced in China are consumed in the US, but the environmental footprint of that production is in the former). But such an approach is constrained by the lack of relevant data (the existing literature on this topic has more narrowly focused specific commodities, on bilateral or regional trade). If these outsourced environmental impacts were incorporated into the appropriate indicators, our findings would likely show that affluent countries have been even less successful at reducing environmental footprints. This phenomenon has important implications for the international distribution of ethical burdens, especially since climate change impacts, pandemic risks, and other global environmental problems will disproportionally affect lower-income countries, which tend to have lower adaptive capacity compared to affluent states [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].

A representative indicator-based approach to assessing sustainability at the national level has been used in other studies, yielding instructive lessons for research and policy [@Holden:2014; @Moran:2008; @Pulsell:2015]. However, we recognize that there are other valid ways to conceptualize sustainability. Modern studies on the relationship between ecological limits, human population, and the economy have a long pedigree [@Ehrlich:1971; @Meadows:1972].  More recently, an influential work by Raworth [-@RaworthKate:2017] postulated 11 "social thresholds" and 7 planetary boundaries to delimit sustainability. This and similar frameworks imply equal weight among chosen thresholds, which mix more subjective indicators (e.g., life satisfaction and democratic quality) with objective measures of wellbeing (e.g., nutrition and healthy life expectancy). Such diverse classification schemes are instructive but may also be over determined: it is conceivable that a different and even longer list of thresholds could be advanced to form an equally relevant sustainability framework. While our three canonical dimensions framework is also imperfect (or perhaps just coarse-grained), its conceptual parsimony accords with long-standing definitions of sustainability.

# Conclusions

Aim (500w | 211w )

Countries around the world are facing a trilemma when it comes to achieving economic prosperity, social equality while reducing environmental impact. While the political strategies to navigate the trilemma space will be context and path dependent, our analysis shows that there are common typologies of how the trilemma has historically evolved. These typologies might help identify groups of countries where common strategies to face the trilemma could work beyond context dependence, bearing in mind that there are no blue print solutions (Ostrom). Whether the patterns of synergies and tradeoffs here identified hold across scales, e.g. looking at finer subnational scales or other units of analysis such as cities or businesses, remains an open question. Most importantly, the causal mechanisms explaining the synergies and tradeoffs between inequality, economic prosperity and the environment are likely to be diverse, rather than single relationships. The foundational role of nature and support system of biodiversity cannot be understated when thinking about sustainable development goals. A myopic focus on prosperity is clearly misplaced. Highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental footrpints. A myopic focus on economic prosperity actually undermines progress towards sustainable development, and the noble ambitions of the international agenda of leaving no one behind.

# Acknowledgements

We would like to thank the conveners and other participants of the Beijer Young Scholars workshops (2016-2018) for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support.  This work was also supported by funding from Formas (grant number 2018-02318).

# References {.unnumbered}

::: {#refs}
:::

\pagebreak

# Supplementary Material {#sec:SM}

```{=tex}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{prosperity_missing_values}
\caption{\textbf{Data quality for prosperity} Different prosperity metrics are compared with respect to their coverage in space (countries) and time (A). Gross national income (GNI) is the metric with higher completion rate (B), with an arbitrary threshold of 30\% missing values, the analysis could include over 100 coutries (C).}
\label{fig:SM_gni_missing}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{inq_missing_values}
\caption{\textbf{Data quality for inequality} Mean Gini income is the most complete statistic of inequality with respect to their coverage in space (countries) and time (A). Alternative metrics such as the Palma index overcome some of the limitations of Gini but have disproportionally more missing values (B). For Gini, the proportion of countries with missing values is above 50\% for any given year. An arbitrary treshold of 30\% threshold for includsion will leave countries like South Africa (country with highers inequality in the world) out of the analysis (D).}
\label{fig:SM_inq_missing}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{ecofp_missing_values}
\caption{\textbf{Data quality for ecological footprint} Coverage of the ecological footprint in space (countries) and time (A) shows that data is relatively complete after mid 1990s (B) An arbitrary threshold of 10 missing values would leave some of the small islands states out of the analysis (C).}
\label{fig:SM_ecofp_missing}
\end{figure*}
```

### 2022-05-10 references.R

@@ -1,0 +1,53 @@
+library(tidyverse)
+library(XML)
+library(RefManageR)
+
+dat <- xmlParse("trilemma_refs.xml")
+
+dat_df <- dat["//record"] |> xmlToDataFrame() |> 
+    as_tibble() |> janitor::clean_names()
+
+dat["//record/ref-type"] |> unlist(recursive = FALSE) 
+l <- dat["//record/contributors"] |> unlist() #|> 
+    # map(function (x) {str_extract(x, "<style*/style>")})
+
+dat_df <- dat_df |> 
+    select(-c(database:foreign_keys), -c(pub_location:num_vols)) |>  #pull(ref_type ) |> unique()
+    mutate(ref_type = case_when(
+        ref_type == 17~ "Article", 
+        ref_type == 13~ "Misc",
+        ref_type == 6~ "Book", 
+        ref_type == 27~ "Report", 
+        ref_type == 59~ "Misc",
+        ref_type == 28~ "Book",
+        ref_type == 19~ "Article",
+        ref_type == 36~ "Misc",
+        ref_type == 43~ "Article",
+    )) |> rename(bibtype = ref_type, author = contributors, title = titles, year = dates, journal = periodical) |> 
+    mutate(institution = case_when(bibtype == "Report" ~ publisher)) 
+
+stops <- str_locate(dat_df$author, "^*,")
+
+dat_df <- dat_df |> 
+    mutate(stop = stops[,1]) 
+dat_df$stop[is.na(dat_df$stop)] <- 5
+
+dat_df <- dat_df |> 
+    mutate(firstauthor = str_trunc(author, width = stop-1 , ellipsis = "")) |> 
+    mutate(firstauthor = str_remove_all(firstauthor, "([:punct:]|[:space:])")) |> 
+    mutate(key = str_c(firstauthor, year, sep = ":")) |> 
+    group_by(key) |> 
+    add_count() |> 
+    mutate(ord = order(key)) |> 
+    mutate(key = case_when(n > 1 ~ paste0(key, letters[ord]), TRUE ~ key)) |> 
+    select(-firstauthor, -n, -ord,-stop) 
+# 
+# dat_df |>
+#     as.BibEntry() |>
+#     WriteBib(file = "references_old.bib")
+
+dat_df |> 
+    select(year, key) |> 
+    print(n=121)
+
+

### 2022-06-23 paper/manuscript.Rmd
incorporating authors comments

@@
 abstract: |
-  Sustainability is often framed as achieving prosperity, equality, and low environmental impact.
+  Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact.
 But what if there is a trilemma that inhibits the joint achievement of these three key goals?
-Here we analyzed the synergies and tradeoffs between these goals by investigating historical data on per-capita gross national income (prosperity), the Gini coefficient for income distribution (inequality), and per-capita ecological footprint (environmental impact).
-We find that no country in the world have successfully solved the trilemma, but there are common typologies that can inform common barriers and strategies to solve it.
-Some countries are trapped on a high inequality space, while prosperous countries can reduce inequality but typically at expenses of high environmental impacts.
-Achieving ambitous sustainable goals will require redistribution of wealth and responsibility for the stewardship of the planet.
+We analyzed synergies and tradeoffs between these dimensions of sustainability by investigating historical data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint.
+We found that no country in the world successfully solved this trilemma, but there are typologies of countries that can help identify common barriers, and strategies to overcome them.
+Countries typically reduce inequality and then increase prosperity and environmental impact.
+Some countries are trapped in a high inequality space, while prosperous countries have reduced inequality only at the expense of increasing environmental impact.
+Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.
 
@@
 Sustainability has often been characterised as having three essential, nested dimensions: economy, society, and the environment.
-More specifically, sustainability has been defined as attaining high material living standards (prosperity), that are also well distributed (equality) and maintainable over the long run (achieved through environmental integrity”) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008].
-Although each is undeniably desirable for development, what if all three goals are not simultaneously achievable?
-It has often been assumed that the ends to which individuals as well as societies aspire inevitably cohere, but *a priori*, there is no reason to believe that diverse goals must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
+More specifically, sustainability has been defined as attaining high material living standards (prosperity) that are well distributed (equality) and maintainable over the long run (achieved through environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008].
+Although each is a desirable goal of development, what if all three are not simultaneously achievable?
+It has often been assumed that the positive ethical ends to which individuals as well as societies aspire can cohere, but *a priori*, there is no reason to believe that diverse goals, however separately desirable, must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
 @Greenspan:2001].
@@
 Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011].
-Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in improving diet, mitigating diseases, and environmental integrity [@Clark:2018].
-Another such “trilemma” that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.
+Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018].
+In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.
 
@@
 If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium].
-And even today, approximately 10% of the world’s population still lives under the international poverty line (less than US\$ 1.90 dollars per day), with a much larger share subsisting above that threshold but still aspiring for higher and more secure living standards
+And even today, approximately 10% of the world’s population still lives under the international poverty line of US\$ 1.90 dollars per day (a derisory standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]), with a much larger share barely subsisting above that threshold and aspiring for higher and more secure living standards
 [@TheWor:2018].
-In a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.
+Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.
 
-At the same time, while many average incomes for countries in what was once called the “Third World” have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
-The widespread feelings of injustice that such unequal distribution arouses has historically been called the "Social Question"[@Faist:2009].
-Over the past decade, inequality has become an ever-urgent issue for policymakers as well as the general public [@Chancel:2020].
-This urgency has been strengthened by a growing body of research demonstrating links between inequality and human wellbeing, health, environmental quality, and economic complexity [@Hamann:2018; @Leach:2018;
+At the same time, while average incomes for many countries in what was once called the "Third World" have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
+The widespread feelings of injustice that such unequal economic distribution arouses has historically been called the "Social Question"[@Faist:2009].
+Over the past decade, inequality has become an increasingly urgent issue for policymakers as well as the general public [@Chancel:2020; @UN2020].
+This urgency has been strengthened by a growing body of research demonstrating links between inequality and economic complexity, human health and wellbeing, and environmental conditions[@Hamann:2018; @Leach:2018;
 @Pikett:2014; @Wilkinson:2009; @Hartmann:2017].
 
-Environmental conditions form the third, and arguably most fundamental, dimension of the sustainability trilemma.
+Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma.
 There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere [@Díaz:2019].
@@
 
-Although all three goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two.
-Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly prioritize prosperity, equality, and the environment [@Sachs:2015].
-In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by a given country and by the world as a whole.
+Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly address prosperity, equality, and the environment [@Sachs:2015].
+However, while all three of these development goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two.
+In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole.
 In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, and if it does exist, where does it occur and why does it occur there?
@@
 consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018].
-We operationalized the trilemma as the movement of countries overtime in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}).
+We operationalized the trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}).
 There is a development synergy if a country moves in desired directions towards more than one goal (e.g.
@@
 pervades the achievement of another (negative correlations).
-It is important to note that in this framing, the sign of the correlation changes depending on the term used: i.e., if one speaks of equity as opposed to inequality, or environmental quality as opposed to environmental footprint.
-In the remainder of this paper we use inequality and environmental footprint because that is how the original data was structured.
-Therefore, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).
+It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e.
+if one speaks of equality as opposed to inequality, or environmental quality as opposed to environmental footprint.
+In the remainder of this paper, we use inequality and environmental footprint because that is how the original data are defined and structured.
+Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).
 
@@
 \centering
-\includegraphics[width = 4in, height = 4in]{conceptual_fig}
-\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality while increasing prosperity and environmental quality.
-There could be synergies and trade-offs in achieving these three objectives.
-We explore trajectory typologies on the trilemma space.}
+\includegraphics[width = 4in, height = 4in]{fig1_trilemma_space}
+\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A).
+There could be synergies and trade-offs in achieving these three objectives (B-D).
+We explore trajectory typologies in the trilemma space.
+53 trajectories from 1995 to 2017 are depicted.
+The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (E), inequality and prosperity (F), and ecological footprint and prosperity (G).
+Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
 \label{fig:trilemma}
@@
 To quantify prosperity, equality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used
-per-capita gross national income (GNI) in US\$ values.GNI is a cognate measure of gross domestic product (GDP) and the two are close in value for most countries.
+per-capita gross national income (GNI) in US\$ values.GNI is related to gross domestic product (GDP) and the two are close in value for most countries.
 A key difference is that GNI accounts for earnings from foreign investments.
 The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source:
-<http://data.worldbank.org>). The threshold for being "prosperous" was   US\$12,746 per capita per year, which is the World Bank’s definition for a high-income country [@TheWor:2018].
+<http://data.worldbank.org>).
+The threshold for being "prosperous" was US\$12,746 per capita per year, which is the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].
 
@@
 3.4 (source: [https://www.wider.unu.edu/data](https://www.wider.unu.edu/data)).
-We used 0.30 as the threshold below which countries are identified as having low inequality, similar to findings from a historical survey by Piketty [@Piketty:2014].
+We used 0.30 as the threshold below which countries are identified as having low inequality, similar to findings from previous historical surveys [@Piketty:2014].
 
 We used per-capita ecological footprint to measure environmental impact.
-Ecological footprint is a holistic estimate of the biophysical resources used by human activity, in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998].
-Its unit is the "global hectare", combining measurements of carbon emissions, built-up area, fishing grounds, croplands, forests, and grazing land into an index of environmental impact.
-In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological “deficit”.
+Ecological footprint is a holistic estimate of the biophysical resources used by human activity, quantified in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998].
+Its unit is the "global hectare", combining measurements of carbon emissions, built-up land, fishing grounds, croplands, forests, and grazing land into an index of environmental impact.
+In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological "deficit".
 Though it is not exhaustive, the ecological footprint provides the most widely-used and data-rich approximation of overall human impact on the environment, and it accounts for the effects of trade on resource consumption by country.
 Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)).
-We used 1.68 global per capita hectares as a threshold for environmental impact, as it reflects the amount of average land needed to satisfy human needs without exceeding the global biocapacity to recover.
-In other words, it is the limit at which current generations’ needs can be satisfied without compromising the ability of future generations to do so.
-Ecological footprint estimates above 1.68 global per capita hectares, therefore, represent unsustainable levels of environmental impact.
+We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed the global biocapacity (i.e., ecological deficit).
+In other words, it is the limit at which current generations' needs can be satisfied without compromising the ability of future generations to satisfy their needs.
 
@@
 
-We used these data to chart the movement of countries across dimensions of prosperity, equality, and environmental integrity over time.
+We used these data to chart the movement of countries across the dimensions of prosperity, equality, and environmental impact over time.
 We further investigated whether these movements trace out typologies of countries in the trilemma space.
@@
 We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz].
-We used principal component analysis as an ordination method for visualization purposes, and vector fitting to show in the principal component space how different features of the trilemma space pull clusters apart.
+We used principal component analysis as an ordination method for visualization, and vector-fitting to show how clusters are pulled apart in the principal component space.
 All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).
 
-An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces is only possible for 53 countries.
-The most incomplete dataset is the inequality data (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
-Based on the quality of the three datasets, we selected time series from 1995-2017 for countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation.
+An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces was only possible for 53 countries.
+The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
+Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation.
 This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemmma space (Fig \ref{fig:cube_time}).
@@
 
-(Aim 1200w | 610w)
-
-```{=tex}
-\begin{figure*}[ht]
-\centering
-\includegraphics[width = 6in, height = 4in]{cube_time}
-\caption{\textbf{Countries trajectories in the trilemma space} 53 trajectories from 1995 to 2017 are depicted for countries with $\leqslant 30\%$ missing values.
-The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (A), inequality and prosperity (B), and ecological footprint and prosperity (C).
-Thresholds that delimit the objective space are depicted as dotted lines.}
-\label{fig:cube_time}
-\end{figure*}
-```
-
-No country in the data simultaneously achieved high prosperity, low inequality, and high environmental standards.
+No country simultaneously achieved high prosperity, low inequality, and low environmental impact.
 This strongly suggests that globally, there is a sustainability trilemma.
-At the country level, there were strong tradeoffs between inequality and ecological footprint (Fig \ref{fig:cube_time}A - as the Gini indicator for income decreases, ecological footprint increases) and ecological footprint and prosperity (Fig \ref{fig:cube_time}C- ecological footprint moves in the same direction as per-capita GNI).
-Additionally, prosperity and inequality (Fig \ref{fig:cube_time}B) were also correlated, with a number of high-income countries (e.g., Sweden, New Zealand) having low levels of inequality in the past, but then showed high variation and a trend toward higher inequality levels in more recent years.
+At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}F- per-capita GNI and ecological footprint move in the same direction).
+Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.
 
 While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typology.
-To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ is the angle in the inequality and prosperity plane while $\theta$ is the angle in the inequality and ecological footprint plane.
-These two angles and the total distance covered in the three-dimensional space are used to uncover country typologies.
+To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane.
+These two angles and the total distance covered in the three-dimensional space are used to uncover typologies in country trajectories.
 
@@
 \includegraphics[width = 6in, height = 4in]{trilema_ensamble}
-\caption{\textbf{Trilemma attractors} }
+\caption{\textbf{Countries' positions and trajectories within the trilemma space} 53 country trajectories from 1995 to 2017 are depicted.
+Colours and contours represent areas of similar values and enclose attractors within peaks.
+The insets of each graph show the direction of movement for countries during the study period.}
 \label{fig:cube_ensamble}
@@
 
-There are certain regions of the trilemma space where countries tend to dwell (Fig \ref{fig:cube_ensamble}).
-The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three clusters of countries.
+There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
+The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters.
 Despite this clustering, countries showed distinctive profiles in terms of their direction of movement in the trilemma space (Fig \ref{fig:cube_ensamble} insets).
@@
 A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995.
-For example, between 1995 and 2017, China's inequality and ecological footprint worsened while GNI per capita grew significantly.
-And while in Ukraine inequality and environmental impact actually fell, prosperity only increased marginally.
+For example, between 1995 and 2017, China's Gini and ecological footprint worsened while GNI per capita grew significantly.
+And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally.
 To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).
 
-The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}).
-First, there was a group of countries trapped in a region of high inequality, mainly in Latin America.
+The trajectories of countries in the trilemma space from 1995 to 2017 converged into three general types (Fig \ref{fig:pca}A).
+First, there was a group of countries "trapped" in a region of high inequality, mainly in Latin America.
 They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$, that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green).
@@
 The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).
-If we ignore temporal dynamics and focus on the mean and standard deviation as summary statistics, we are able to include more countries (N=140).
-Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and the environment.
+If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma dimensions as summary statistics, we are able to include more countries (N=140).
+Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and environmental impact.
 The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.
@@
 \centering
-\includegraphics[width = 6in, height = 3in]{pca_trilemma}
+\includegraphics[width = 6in, height = 3in]{pca_maps}
 \caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies.
@@
 It complements the dynamic analysis with one summary statistic over time for 140 countries (B).
-}
+Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
 \label{fig:pca}
@@
 
-(Aim 1000 | 1472)
+Our analysis found that decreasing inequality and ecological impact pulled countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time.
+The historical data shows that countries typically reduce inequality and then increase environmental impact, not the other way around (Fig \ref{fig:cube_ensamble}A).
+Similarly, countries tend to reduce inequality first, then increase prosperity (Fig \ref{fig:cube_ensamble}B).
+Importantly, many middle-income countries have levels of inequality similar to high-income countries, but at lower levels of environmental impact and prosperity (Fig \ref{fig:cube_ensamble}B, C).
+This means that countries can achieve relatively low levels of inequality while attaining and preserving environmental integrity, but our findings show that higher prosperity often leads to higher environmental impact.
+On the other hand, our results also indicate that some countries struggled to escape areas in the trilemma space characterized by high levels of inequality during the period under consideration.
+However, it is important to note that past trajectories do not necessarily determine plausible future pathways.
 
-Our analyses found that inequality and ecological footprint goals pull countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time.
-This suggests the existence of regions within the trilemma space that are hard to escape, characterized by high levels of inequality.
-The historical data shows that countries typically reduce inequality and then increase ecological footprint, not the other way around (Fig \ref{fig:cube_ensamble}A).
-Importantly, many middle-income countries have similar levels of inequality as high-income ones, but at lower levels of ecological footprint and per-capita GNI (Fig \ref{fig:cube_ensamble}B, C).
-This means that countries can achieve low levels of inequality while attaining and preserving high environmental quality.
-Historical evidence shows that higher prosperity often leads to higher ecological footprints.
-Nonetheless, it is important to note that past trajectories do not necessarily determine plausible future pathways in the trilemma space.
-
-```{=tex}
-\begin{figure*}[ht]
-\centering
-\includegraphics[width = 6in, height = 1.5in]{maps_trilemma}
-\caption{\textbf{Countries mean performance on trilemma space} Mean values of ecological footprint (A), inequality (B), and gross national income (C). }
-\label{fig:map}
-\end{figure*}
-```
-The potential mechanisms of the tradeoffs and synergies between economic, social, and environmental goals remain open questions.
+The potential mechanisms of tradeoffs and synergies remain an important research frontier.
 Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC).
 In this framing, the former increases as the later grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955].
-Since this relationship was first posited in 1955, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020].
+Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020].
 This has made the identification of mechanisms difficult.
-Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through systemic regularities of the global economy or proactive policies by national governments [@Pikett:2014].
-In our analysis of the trilemma space, we did not find evidence for a KC-like relationship.
+Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014].
+Our analysis found no evidence for a KC-like relationship, instead finding that increasing inequality tended to precede growing prosperity.
 
-In terms of the economy-environment interface, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship and implicit explanation: countries increase their environmental footprints as incomes rise (which in modern history has been driven by industrializatoin and urbanization); but after reaching an inflection point of affluence, the footprints of an increasingly rich society decline (Barbier 1997).
+In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline (Barbier 1997).
 As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not (Dasgupta et al.
@@
 Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns (Mäler 2013).
-But what is clear from the wealth of diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.
+But what is clear from diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.
 
-Other mechanisms and pathways that do not posit linear or quadratic relationships have been identified as driving dynamics between economic, social and environmental dimensions.
-For instance, countries with high per-capita biocapacity tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions [@Weinzett:2013].
+Other mechanisms and pathways that do not posit linear or quadratic relationships have been identified as driving interactions between the economic, social and environmental dimensions of sustainability.
+For instance, Weinzettel et al.
+[@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
 Similarly, population, environmental carrying capacity, and pollution exhibit mutually-reinforcing feedbacks across countries [@LaTorr:2019].
@@
 
-Additionally, structural forces in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens.
-Environmental impacts can be “outsourced”, as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses (Mäler 2013).
+Additionally, structural inequities in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens.
+Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses (Mäler 2013).
 For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013].
-Additionally, it has been shown that international trade has significantly influenced global land use patterns, worsening natural resource exploitation and biodiversity loss [@Creutzig:2019].
-Thus telecoupling and resource consumption from far away places is a mechanism that can maintain countries in undesirable regions of the trilemma space.
+Additionally, it has been shown that international trade has significantly influenced global land use patterns, worsening natural resource exploitation and biodiversity loss[@Creutzig:2019].
+This "telecoupling" of unsustainable resource use between far away places may be a mechanism that can obfuscate real trajectories in the trilemma space, undermining the ability to reach all three goals at the global level by entrapping certain countries in undesirable regions.
 
-While no country was found to have eluded the sustainability trilemma, there were states, forming clusters, in which tradeoffs and synergies between economic, social and environmental goals moved in distinctively different directions.
-These dynamics further highlight the inadequacy of reductionistic, often mono-causal, explanations - such as those found in many KC and EKC studies - and the need to incorporate historical, geographic, and structural factors.
-For instance, in Ukraine, the existence of (relatively) low inequality and environmental footprint alongside low prosperity seemed to conform to the KC and EKC relationships.
-However, inequality levels in the Philippines, which had a similar average income, and Mexico, which had a similar per-capita ecological footprint, were extremely high.
-Additionally, Ukraine’s trajectory through the trilemma space was far from smooth: over the study period, average incomes rose and fell in dramatic downswings due to corruption (endemic and structural across post-Soviet Union countries) and war, among other historical factors.
+While no country was found to have eluded the sustainability trilemma, there were clusters that moved in distinctively different directions with respect to tradeoffs and synergies between the three goals.
+These dynamics further highlight the inadequacy of reductionistic, often mono-causal, explanations - such as those found in many KC and EKC studies - and the need to incorporate historical, geographic, and institutional factors.
+For instance, at the end of the study period, Ukraine had (relatively) low levels of inequality and environmental impact.
+However, the country’s  trajectory through the trilemma space was far from smooth: over the study period, indicators  rose and fell in dramatic downswings due to market shocks and corruption (endemic across post-Soviet countries throughout the 1990s) as well as  dislocations caused by war.
+In fact, it was only after the Russo-Ukrainian conflict began in 2013 that Ukraine’s ecological footprint fell below its biocapacity (Fig.
+\ref{fig:S4}).
 
-The clustering of countries in the trilemma space also shows the importance of geographic factors, especially territorial features and natural resource endowments, to sustainable development.
-Countries that performed relatively well along all three dimensions tended to be advanced Western economies, particularly along the North Atlantic rim (e.g., Canada, Sweden) and in the Antipodes (i.e., New Zealand and Australia).
-These countries, in addition to benefiting from the legacies of early industrialization and imperialism, also had a much higher ratio of space and resources to people, as reflected for instance in per-capita biocapacity.
+The clustering also shows the importance of geographic factors, including the natural resource endowment factor mentioned earlier.
+Countries that performed relatively well along all three dimensions tended to be advanced Western economies, particularly along the North Atlantic rim (e.g., Canada, Sweden).
+These countries had a much higher ratio of space and resources to people, as reflected in per-capita biocapacity.
 Finally, the structure of the global economy is also a powerful factor in explaining international differences in the trilemma space.
-China, the rapid rise of both prosperity, inequality, and environmental footprint were significantly driven by its integration into international networks of commerce and finance (most notably its accession to the World Trade Organization in 2002).
-Rapid economic growth was driven by the emergence of a resource-intensive manufacturing sector - with China becoming the so-called “factory of the world”.
-At the same time, studies show that the surge of economic inequality in China was due in large part to geographic differences: coastal areas were the manufacturing base and export hubs, with high-value added contributions, while the inland areas supplied mostly raw materials, intermediate goods, and cheap labour.
+In China, the rapid rise of both prosperity, inequality, and environmental impact were significantly driven by its integration into international commercial networks (Fig.
+\ref{fig:S4}).
+Rapid economic growth was driven by a resource-intensive export sector while rising income inequality was due in large part to geographic differences: urban coastal areas were the affluent export hubs while more rural inland areas supplied mostly raw materials, intermediate goods, and cheap labour [@xie2014income; @qi2013interpreting].
 
-Although we believe our analysis of the sustainability trilemma identified important patterns and drivers of international development (as well as lacunae in the existing literature), it still faced limitations of data and method.
-First, the offshoring of environmental degradation discussed above is also not completely captured in standard ecological footprint measurements [@VandenBergh:2012].
-A bottom-up or consumption-based approach to measuring environmental impacts – as opposed to the more mainstream top-down or production-based approach – may help internalize these externalities (e.g., resource-intensive goods produced in China are consumed in the US, but the environmental footprint of that production is in the former).
-But such an approach is constrained by the lack of relevant data (the existing literature on this topic has more narrowly focused specific commodities, on bilateral or regional trade).
-If these outsourced environmental impacts were incorporated into the appropriate indicators, our findings would likely show that affluent countries have been even less successful at reducing environmental footprints.
-This phenomenon has important implications for the international distribution of ethical burdens, especially since climate change impacts, pandemic risks, and other global environmental problems will disproportionally affect lower-income countries, which tend to have lower adaptive capacity compared to affluent states [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].
+Although our analysis identified important patterns and drivers of international development (as well as lacunae in the existing literature), it still faced practical limitations.
+First, the offshoring of environmental degradation discussed above is not completely captured in standard ecological footprint measurements [@VandenBergh:2012].
+A "bottom-up" or consumption-based approach to measuring environmental impacts – as opposed to the more mainstream "top-down" or production-based approach – may help internalize these externalities (e.g., resource-intensive goods produced in China are consumed in the US, but the environmental footprint of that production is attributed to the former).
+But such an approach is constrained by the lack of relevant data (the existing literature on this topic has more narrowly focused on specific commodities and on bilateral or regional trade).
+If these outsourced environmental impacts were incorporated into the appropriate indicators, our findings would likely show that affluent countries have been even less successful at reducing their ecological footprints.
+This phenomenon has important implications for the international distribution of ethical burdens, especially since climate change impacts, pandemic risks, and other global environmental problems will disproportionally affect lower-income countries, which tend to have lower adaptive capacity [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].
 
@@
 More recently, an influential work by Raworth [-@RaworthKate:2017] postulated 11 "social thresholds" and 7 planetary boundaries to delimit sustainability.
-This and similar frameworks imply equal weight among chosen thresholds, which mix more subjective indicators (e.g., life satisfaction and democratic quality) with objective measures of wellbeing (e.g., nutrition and healthy life expectancy).
-Such diverse classification schemes are instructive but may also be over determined: it is conceivable that a different and even longer list of thresholds could be advanced to form an equally relevant sustainability framework.
-While our three canonical dimensions framework is also imperfect (or perhaps just coarse-grained), its conceptual parsimony accords with long-standing definitions of sustainability.
+This and similar frameworks imply equal weight among chosen thresholds, which mix more subjective indicators (e.g., life satisfaction and democratic quality) with objective measures of wellbeing (e.g., nutrition and life expectancy).
+Such diverse classification schemes are valuable but may also be overdetermined: it is conceivable that a different and even longer list of thresholds could be advanced to form an equally relevant sustainability framework.
+In addition, there have been calls for global frameworks that move beyond the artificial separation of social and ecological goals towards more integrated social-ecological targets for sustainable development, to better account for the foundational role played by nature [@reyers2020global].
+While acknowledging that our three canonical dimensions framework is imperfect (or perhaps just coarse-grained), we believe its conceptual parsimony nonetheless reflects long-standing definitions of sustainability.
 
@@
 
-Aim (500w | 211w )
+Our analysis found that countries around the world face a trilemma when it comes to achieving economic prosperity and social equality while reducing environmental impact.
+While national strategies to navigate the trilemma space will be context- and path-dependent, our findings show that there are common typologies of how countries have historically evolved in the trilemma space.
+These typologies may help identify country groupings for which common development strategies address the challenges of  the sustainability trilemma.
+Nonetheless, it is important to recognize the differing geographic, historical, and institutional constraints facing countries, and that when it comes to the multi-dimensional challenges of sustainable development, there are no "one-size-fits-all" solutions [@Ostrom:2007; @ostrom2010moving; @perrings2007future].
 
-Countries around the world are facing a trilemma when it comes to achieving economic prosperity, social equality while reducing environmental impact.
-While the political strategies to navigate the trilemma space will be context and path dependent, our analysis shows that there are common typologies of how the trilemma has historically evolved.
-These typologies might help identify groups of countries where common strategies to face the trilemma could work beyond context dependence, bearing in mind that there are no blue print solutions (Ostrom).
-Whether the patterns of synergies and tradeoffs here identified hold across scales, e.g.
-looking at finer subnational scales or other units of analysis such as cities or businesses, remains an open question.
-Most importantly, the causal mechanisms explaining the synergies and tradeoffs between inequality, economic prosperity and the environment are likely to be diverse, rather than single relationships.
-The foundational role of nature and support system of biodiversity cannot be understated when thinking about sustainable development goals.
-A myopic focus on prosperity is clearly misplaced.
-Highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental footrpints.
-A myopic focus on economic prosperity actually undermines progress towards sustainable development, and the noble ambitions of the international agenda of leaving no one behind.
+Whether the patterns of synergies and tradeoffs identified in this study hold across scales (e.g., looking at finer subnational scales or other analytic units such as cities or corporations), remains an open question.
+Most importantly, the causal mechanisms explaining the synergies and tradeoffs between inequality, prosperity, and environmental impact are likely to be diverse, rather than mono-causal relationships.
+The foundational role of nature, which provides the basic life-support systems for humanity, cannot be understated when thinking about sustainable development goals.
+As our findings show, a myopic focus on prosperity is clearly misplaced - neither KC or EKC relationships were found for countries over the twenty-year study period.
+Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts.
+An overemphasis on  economic prosperity can undermine progress towards the noble ambitions of leaving no country or person behind in the pursuit of global sustainability.
 
@@
 ```
+
+```{=tex}
+\begin{figure*}[ht]
+\centering
+\includegraphics[width = 6in, height = 6in]{FigS4}
+\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the “trilemma space” formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
+\label{fig:S4}
+\end{figure*}
+```

### 2022-07-01 paper/manuscript.Rmd
final edits for submission

@@
 title: "Global tradeoffs between prosperity, inequality, and the environment"
-# author: |
+author: |
+    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^, Emilie Lindkvist^2^, Maike Hamann^6^, Jiangxiao Qiu^7^, 
+    | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
+    | \small
+    | \footnotesize ^1^The Natural Capital Project and Stanford Center on China's Economy and Institutions, Freeman Spogli 
+    | \footnotesize Institute for International Studies, Stanford University
+    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
+    | \footnotesize ^3^Future Earth
+    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
+    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
+    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
+    | \footnotesize ^7^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
+    | \footnotesize ^8^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
+    | \footnotesize ^9^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
+    | \footnotesize ^§^These authors made equal contributions
 #   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
 abstract: |
+
   Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact.
@@
 Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.
-
 output:
@@
 
+## word counts
+## Abstract 150
+## Intro 697
+## Method 957
+## Results 621
+## Discussion 1434
+## Conclusion 267
+## Total main 3976 + 150 abstract
+## 
+## Targets: 
+# 
+# - NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
+# 70max
+# 
+# - WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
+# abstract
 ```
 
-Targets:
-
-- NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
-70max
-
-- WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
-abstract
-
-Current: 3957w
-
 # Introduction
-
-(Aim 1000w | Current 671w)
 
@@
 Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011].
-Additionally, researchers suggest that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018].
+Additionally, research suggests that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018].
 In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.
@@
 [@TheWor:2018].
-Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world.
+Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].
 
@@
 
-Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma.
-There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere [@Díaz:2019].
+Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic].
+Human well-being intimately depends on it and it forms the precondition for economic development.
+There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere, our home [@Díaz:2019; @folke2021our].
 Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
 [@Díaz:2019].
-At the same time, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017].
+Simultaneously, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017].
 Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].
@@
 In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole.
-In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, and if it does exist, where does it occur and why does it occur there?
+In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, are there inherent tradeoffs between these goals, and if it does exist, where does it occur and why does it occur there?
 
 # Methods
-
-(Aim 800w | current 972w)
 
@@
 There is a development synergy if a country moves in desired directions towards more than one goal (e.g.
-reducing inequality and increasing prosperity) There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another.
+reducing inequality and increasing prosperity).
+There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another.
 Therefore, synergies are movements along
@@
 while tradeoffs are represented by the (red) axis where the achievement of one
-pervades the achievement of another (negative correlations).
+hinders the achievement of another (negative correlations).
 It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e.
@@
 \centering
-\includegraphics[width = 4in, height = 4in]{fig1_trilemma_space}
+\includegraphics[width = 6in, height = 4in]{fig1_trilemma_space}
 \caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A).
 There could be synergies and trade-offs in achieving these three objectives (B-D).
-We explore trajectory typologies in the trilemma space.
+We explore trajectory typologies in the trilemma space (E-G).
 53 trajectories from 1995 to 2017 are depicted.
@@
 To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.
-We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters to fit.
+We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters.
 We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz].
@@
 
-An analysis of the missing values of the publicly available data shows that a assessing the change over time in the trilemma spaces was only possible for 53 countries.
+An analysis of the missing values of the publicly available data shows that assessing the change over time in the trilemma spaces was only possible for 53 countries.
 The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
 Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation.
-This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemmma space (Fig \ref{fig:cube_time}).
+This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).
 
@@
 This strongly suggests that globally, there is a sustainability trilemma.
-At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}F- per-capita GNI and ecological footprint move in the same direction).
+At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction).
 Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.
@@
 ```{=tex}
-\begin{figure*}[ht]
+\begin{figure*}[htp]
 \centering
-\includegraphics[width = 6in, height = 3in]{pca_maps}
+\includegraphics[width = 6.5in, height = 5in]{pca_maps}
 \caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies.
-3 clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A).
-2 clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness.
+Three clusters are identified when looking at the historical trajectory of countries (N=53) from 1995-2017 (A).
+Two clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness.
 It complements the dynamic analysis with one summary statistic over time for 140 countries (B).
+For both A and B, red arrows are vectors fitted to the data that indicate the directions along which countries and clusters differ from each other.
 Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
@@
 Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC).
-In this framing, the former increases as the later grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955].
+In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955].
 Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020].
-This has made the identification of mechanisms difficult.
+This has made the identification of mechanism difficult.
 Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014].
@@
 
-In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline (Barbier 1997).
-As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not (Dasgupta et al.
-2002; Stern 2004).
-In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC.
-Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns (Mäler 2013).
+In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline [@barbier1997economic].
+As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not [@dasgupta2002confronting; @stern2004rise].
+In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC, i.e.
+an inflection point in the prosperity-environment plane (Fig \ref{fig:cube_ensamble}).
+Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325].
 But what is clear from diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.
@@
 Additionally, structural inequities in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens.
-Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses (Mäler 2013).
+Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses [@MALER201325].
 For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013].
@@
 We would like to thank the conveners and other participants of the Beijer Young Scholars workshops (2016-2018) for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support.
-This work was also supported by funding from Formas (grant number 2018-02318).
+This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.
+
+**Author contributions:**  Conceptualization was a team effort, TW and JCR produced the first manuscript, TW, JCR and JQ analyized the data. All authors read and edit the paper.
 
@@
 \centering
-\includegraphics[width = 6in, height = 6in]{FigS4}
-\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the “trilemma space” formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
+\includegraphics[width = 6in, height = 5in]{FigS4}
+\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the "trilemma space" formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
 \label{fig:S4}

### 2023-01-25 paper/manuscript.Rmd
prep new submission, include Tong re-write

@@
 #   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
-abstract: |
 
-  Sustainability is often framed as achieving prosperity, equality, and environmental integrity: in other words, high living standards, evenly-distributed, and low-impact.
-But what if there is a trilemma that inhibits the joint achievement of these three key goals?
-We analyzed synergies and tradeoffs between these dimensions of sustainability by investigating historical data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint.
-We found that no country in the world successfully solved this trilemma, but there are typologies of countries that can help identify common barriers, and strategies to overcome them.
-Countries typically reduce inequality and then increase prosperity and environmental impact.
-Some countries are trapped in a high inequality space, while prosperous countries have reduced inequality only at the expense of increasing environmental impact.
-Escaping this trilemma to achieve ambitious sustainability goals will likely require economic redistribution and more urgent collective action for planetary stewardship.
 output:
@@
     toc: no
-csl: /Users/juanrocha/Documents/styles/nature.csl
+csl: /Users/juanrocha/Documents/styles/apa.csl
 bibliography: references_ei.bib
@@
 - \linenumbers
+- \usepackage{setspace}
 ---
@@
 
+\doublespacing
+
+**Author contributions:**  Conceptualization was a team effort, TW and JCR produced the first manuscript, TW, JCR and JQ analyzed the data.
+All authors contributed to the writing and editing of the paper.
+
+**Declaration of interests:** None
+
+**Correspondence:** juan.rocha@su.se | +46730747766
+
+\pagebreak
+
+# Highlights
+
+- Historical data reveals that no country in the world has reduced inequality and environmental footprint simultanously
+- A trilemma exist where countries need to trade-off progress on social and environmental goals
+- An inequality trap seems to exist particularly acute in Latinamerica
+- Countries trajectories on the trilemma space suggest that in the future social progress can be achieved without compromising the biosphere
+
+\pagebreak
+# Acknowledgements
+
+We would like to thank the conveners and other participants of the 2016-2018 Beijer Young Scholars workshops for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support.
+This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.
+
+\pagebreak
+
+# Global tradeoffs between prosperity, inequality, and the environment
+
+\begin{abstract}
+
+  Sustainability has often been defined as jointly achieving prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact.
+But what if there is a “sustainability trilemma” that inhibits the simultaneous achievement of these three developmental goals?
+We analyzed synergies and tradeoffs between these three key dimensions of sustainability using data on per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017.
+Over these two decades, we found that no country in the world overcame the hypothesized trilemma.
+Nonetheless,the trajectories of  countries in the three-dimensional space formed by these key indicators revealed  typologies that identified  common development barriers as well as potential strategies to overcome them.
+Historically, countries tended to decrease in income inequality before growing in prosperity and environmental impact.
+Some countries were trapped in a space of high inequality, while prosperous countries lowered inequality only at the expense of increasing environmental impact.
+Achieving  greater synergy at the nexus of these three sustainability dimensions will likely require countries - especially those that have already become prosperous - to prioritize economic redistribution and environmental stewardship alongside or even before material growth.
+At the global level, structural inequities in the world economy as well as historical and geographical factors suggest the need for more urgent collective action, especially in the face of the Anthropocene’s worsening risks.
+Without such concerted efforts, the sustainable development of individual countries, and of the world, may continue to be stymied by this trilemma.
+
+\end{abstract}
+
 # Introduction
 
-Sustainability has often been characterised as having three essential, nested dimensions: economy, society, and the environment.
-More specifically, sustainability has been defined as attaining high material living standards (prosperity) that are well distributed (equality) and maintainable over the long run (achieved through environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008].
-Although each is a desirable goal of development, what if all three are not simultaneously achievable?
-It has often been assumed that the positive ethical ends to which individuals as well as societies aspire can cohere, but *a priori*, there is no reason to believe that diverse goals, however separately desirable, must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007;
-@Greenspan:2001].
+Sustainability has often been characterised as having three essential and nested dimensions: economy, society, and the environment.
+More specifically, this tripartite framework  has been defined as development that attains high material living standards (prosperity) that are well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008].
+Although each goal is desirable in its own right, the question nonetheless needs to be asked: What if all three are not simultaneously achievable?
+It has often been assumed that the positive ends to which individuals as well as societies aspire can cohere, but *a priori*, there is no reason to believe that diverse goals must be compatible (across scales) or scalable (without upper bounds) [@Gray:2007; @Greenspan:2001].
 
 In the literature on development, such macro-level incompatibilities are not novel.
-For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination
-[@RodrikD:2011].
+For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination [@RodrikD:2011].
 Using resources efficiently, fairly and with maximum freedom has also been proposed as an unattainable triad [@Coyle:2011].
-Additionally, research suggests that based on international trends, there may be persisting tradeoffs between changes in dietary improvement, disease mitigation, and environmental integrity [@Clark:2018].
-In this light, we argue that another such "trilemma" that countries may face in the Anthropocene is the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.
+Additionally, based on international trends, research suggests that there may be persisting tradeoffs between dietary improvement, disease mitigation, and environmental integrity [@Clark:2018].
+In this light, we argue that there may be another such "trilemma" facing countries in the Anthropocene: the difficulty - or perhaps impossibility - of being prosperous and equal while achieving or maintaining low environmental impact.
 
-For most of human history, scarcity has been a ubiquitous condition of life: material want, not sufficiency, was the lot of most people [@Maddison:2007;
-@Harari:2014].
+For most of human history, material want, not sufficiency, was the norm for most people regardless of county [@Maddison:2007; @Harari:2014].
 Although global prosperity has increased in recent generations, progress has been uneven.
 If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium].
-And even today, approximately 10% of the world’s population still lives under the international poverty line of US\$ 1.90 dollars per day (a derisory standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]), with a much larger share barely subsisting above that threshold and aspiring for higher and more secure living standards
-[@TheWor:2018].
-Furthermore, in a rise of global poverty unprecedented since the end of World War II, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already derisory line, underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].
+And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]).
+Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and understandably aspires for higher and more secure living standards [@TheWor:2018].
+Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an increase rise of global poverty unprecedented since the end of World War II and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].
 
-At the same time, while average incomes for many countries in what was once called the "Third World" have progressively converged with those of advanced, mostly Western, economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
-The widespread feelings of injustice that such unequal economic distribution arouses has historically been called the "Social Question"[@Faist:2009].
-Over the past decade, inequality has become an increasingly urgent issue for policymakers as well as the general public [@Chancel:2020; @UN2020].
-This urgency has been strengthened by a growing body of research demonstrating links between inequality and economic complexity, human health and wellbeing, and environmental conditions[@Hamann:2018; @Leach:2018;
-@Pikett:2014; @Wilkinson:2009; @Hartmann:2017].
+At the same time, while average incomes for certain countries in what was once called the "Third World" began to  converge with those of the advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
+The widespread feelings of injustice that the increasingly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009].
+Over the past decade, inequality has once again become an increasingly urgent issue for policymakers as well as the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020].
+These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks[@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017].
+Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened income and wealth inequalities within many countries, as well as widening the development trajectories between high- and low-income economies [@WorldBank:2022].
 
-Indeed, environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic].
-Human well-being intimately depends on it and it forms the precondition for economic development.
-There is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere, our home [@Díaz:2019; @folke2021our].
-Human activities have become the primary drivers of biodiversity loss, land degradation, and climate change, among other environmental impacts
-[@Díaz:2019].
-Simultaneously, rapid population and economic growth have exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017].
-Global development in the 21st Century requires the protection and restoration of ecosystems as well as the mitigation of catastrophic climate change [@Millenniu:2005; @IPCC:2021].
+Environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic].
+Interdisciplinary research in sustainability science confirms that human well-being and economic development depend on the resilience and abundance of nature.
+At the same time, there is now a scientific consensus that human activities have a wide and deleterious impact on the biosphere, which constitutes the habitative upper bound and common denominator for our species (notwithstanding fanciful visions of extraterrestrial expansion)  [@Díaz:2019; @folke2021our].
+Anthropogenic drivers  have become primarily responsible for global biodiversity loss, land degradation, and climate change, among other environmental impacts [@Díaz:2019].
+Indeed, humanity  has already exceeded a number of the biophysical boundaries that delimit a “safe operating space” for humanity [@Rockström:2009; @RaworthKate:2017].
+Global development in the 21st Century therefore urgently requires more sensible stewardship of the biosphere, from ecosystems to the climate [@Millenniu:2005; @IPCC:2021].
 
-Recognizing the important nexus formed by economic, social, and environmental challenges, the United Nations’ Sustainable Development Goals explicitly address prosperity, equality, and the environment [@Sachs:2015].
-However, while all three of these development goals are necessary, the pursuit of any one may either help or hinder the pursuit of the other two.
+The United Nations’ Sustainable Development Goals explicitly recognize the important nexus formed by prosperity, equality, and the environment  [@Sachs:2015].
+And in its report charting the path towards a sustainable recovery in the wake of COVID-19, the World Bank also identified the need to achieve “shared prosperity” by reducing inequality and accounting for the climate crisis and other environmental challenges [@WorldBank:2022].
+Nonetheless, while all three of these development goals are seen as necessary, it remains for empirical investigation to determine whether the pursuit of any one helps or hinders the pursuit of the other two.
 In this study, we ask whether there is evidence – and rationale – that all three goals can be jointly achieved, by any given country and by the world as a whole.
-In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist, are there inherent tradeoffs between these goals, and if it does exist, where does it occur and why does it occur there?
+In other words, this study attempts to answer the questions: To what extent does the sustainability trilemma exist and are there inherent tradeoffs between these goals?
+If a sustainability trilemma does exist, where does it occur and why does it occur there?
 
@@
 
-To answer these questions, we examined the historical performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}).
-We chose the country scale due to practical considerations
-of data availability, but also on the recognition that many of the most
-consequential policies affecting sustainability are made and implemented at this level [@Bryan:2018].
-We operationalized the trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, inequality, and environmental footprint  (Fig \ref{fig:trilemma}).
-There is a development synergy if a country moves in desired directions towards more than one goal (e.g.
-reducing inequality and increasing prosperity).
+To answer these research questions, we examined the performance of countries over the past quarter-century using publicly available data on the three dimensions of the sustainability trilemma described above (Table \ref{tab1}).
+We chose the country level based on the fact that many of the most consequential policies, plans, and institutions affecting sustainability are made by national governments [@Bryan:2018], and due to practical considerations of data availability.
+We operationalized the sustainability trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, (in)equality, and environmental integrity (alternatively, impact)  (Fig \ref{fig:trilemma}).
+There is a development synergy if a country moves in desired directions towards more than one goal (e.g., reducing inequality and increasing prosperity).
 There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another.
-Therefore, synergies are movements along
-the (blue) diagonal where the achievement of two goals can be positively correlated;
-while tradeoffs are represented by the (red) axis where the achievement of one
-hinders the achievement of another (negative correlations).
-It is important to note that in this framing, the sign of the correlation changes depending on the term used, i.e.
-if one speaks of equality as opposed to inequality, or environmental quality as opposed to environmental footprint.
-In the remainder of this paper, we use inequality and environmental footprint because that is how the original data are defined and structured.
+Therefore, synergies between two goals are movements along the (blue) diagonal (positively correlated), while tradeoffs are represented by the (red) axis, where the achievement of one hinders the achievement of another (negatively correlated).
+It is important to note that in this framing, the sign of the correlation changes depending on the term used.
+For instance, if one speaks of equality as opposed to inequality, or environmental integrity as opposed to environmental impact.
+In the remainder of this paper, we use inequality and environmental impact (proxied by ecological footprint) because this is how the original data was defined and structured.
 Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).
@@
 
-To quantify prosperity, equality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used
-per-capita gross national income (GNI) in US\$ values.GNI is related to gross domestic product (GDP) and the two are close in value for most countries.
+To quantify prosperity, inequality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}).
+For prosperity, we used per-capita gross national income (GNI) in US\$ values.
+GNI is related to gross domestic product (GDP) and the two are close in value for most countries.
 A key difference is that GNI accounts for earnings from foreign investments.
-The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source:
-<http://data.worldbank.org>).
-The threshold for being "prosperous" was US\$12,746 per capita per year, which is the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].
+The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source: <http://data.worldbank.org>).
+The threshold for being "prosperous" was US\$12,746 per capita per year, which was the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].
 
@@
 Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)).
-We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed the global biocapacity (i.e., ecological deficit).
+We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed global biocapacity (i.e., ecological deficit).
 In other words, it is the limit at which current generations' needs can be satisfied without compromising the ability of future generations to satisfy their needs.
@@
 
-We used these data to chart the movement of countries across the dimensions of prosperity, equality, and environmental impact over time.
-We further investigated whether these movements trace out typologies of countries in the trilemma space.
+We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time.
+We further investigated whether these movements trace out typologies of countries within the trilemma space.
 To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.
-We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to test the optimal number of clusters.
+We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters).
 We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz].
-We used principal component analysis as an ordination method for visualization, and vector-fitting to show how clusters are pulled apart in the principal component space.
-All data used are publicly available, and the code used for the analysis is available at: [https://github.com/juanrocha/trilemma](https://github.com/juanrocha/trilemma).
+We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible).
+All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.
 
-An analysis of the missing values of the publicly available data shows that assessing the change over time in the trilemma spaces was only possible for 53 countries.
-The most incomplete dataset is the inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
-Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values, and inputed them with a linear interpolation.
+An analysis of the missing values in the publicly available data shows that assessing changes over time in the trilemma space was only possible for 53 countries.
+The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
+Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values and then imputed them with a linear interpolation.
 This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).
@@
 This strongly suggests that globally, there is a sustainability trilemma.
-At the country level, there were strong tradeoffs between inequality and environmental impact (Fig \ref{fig:trilemma}E - as the Gini indicator for income decreases, ecological footprint increases) as well as increasing prosperity and decreasing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction).
-Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but show high variation and a trend towards greater income inequality in more recent years.
+At the country level, there were strong tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction).
+Results show synergies between increasing prosperity and reducing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden, New Zealand) exhibited low levels of inequality in the past, but also showed high variation and a trend towards greater income inequality in more recent years.
 
-While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typology.
-To further investigate individual trajectories, we calculated the displacement on the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane.
-These two angles and the total distance covered in the three-dimensional space are used to uncover typologies in country trajectories.
+While these globally-aggregated patterns provide insight into international variations with respect to key dimensions of sustainability, they do not necessarily reveal the diversity of individual country trajectories, nor their typologies.
+To further investigate individual trajectories, we calculated the displacement within the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane.
+These two angles and the total distance covered in the three-dimensional space help to identify typologies in country trajectories.
 
@@
 There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
-The inequality and ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity and ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality and prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters.
-Despite this clustering, countries showed distinctive profiles in terms of their direction of movement in the trilemma space (Fig \ref{fig:cube_ensamble} insets).
+The inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters.
+Despite this clustering, countries showed distinctive profiles in terms of their direction of movement (Fig \ref{fig:cube_ensamble} insets).
 Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started.
 A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995.
-For example, between 1995 and 2017, China's Gini and ecological footprint worsened while GNI per capita grew significantly.
+For example, between 1995 and 2017, China's Gini and ecological footprint worsened while per-capita GNI grew significantly.
 And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally.
@@
 First, there was a group of countries "trapped" in a region of high inequality, mainly in Latin America.
-They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$, that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green).
-The second group showed large displacements within the trilemma space but with no definite direction, characterized by particularly high variations in Gini and per-capita GNI (Fig \ref{fig:pca}A, in yellow).
-The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).
+They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations on the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green).
+The second group showed large displacements within the trilemma space but with no definite direction, and was characterized by particularly high variations in its Gini coefficient for income and per-capita GNI (Fig \ref{fig:pca}A, in yellow).
+The third group was composed of countries with relatively higher prosperity and ecological footprint; their movements in the trilemma space were relatively well-defined, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).
 If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma dimensions as summary statistics, we are able to include more countries (N=140).
-Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between inequality, prosperity, and environmental impact.
+Consequently, this yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact.
 The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies, while the second (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia.
@@
 
-Our analysis found that decreasing inequality and ecological impact pulled countries in opposite directions (Fig \ref{fig:pca}) providing strong evidence for the difficulty of achieving both goals at the same time.
-The historical data shows that countries typically reduce inequality and then increase environmental impact, not the other way around (Fig \ref{fig:cube_ensamble}A).
-Similarly, countries tend to reduce inequality first, then increase prosperity (Fig \ref{fig:cube_ensamble}B).
-Importantly, many middle-income countries have levels of inequality similar to high-income countries, but at lower levels of environmental impact and prosperity (Fig \ref{fig:cube_ensamble}B, C).
-This means that countries can achieve relatively low levels of inequality while attaining and preserving environmental integrity, but our findings show that higher prosperity often leads to higher environmental impact.
-On the other hand, our results also indicate that some countries struggled to escape areas in the trilemma space characterized by high levels of inequality during the period under consideration.
-However, it is important to note that past trajectories do not necessarily determine plausible future pathways.
+Overall, our analysis found that decreasing inequality and decreasing environmental impact pulled countries in opposite directions (Fig \ref{fig:pca}), providing strong evidence for the difficulty of achieving both goals at the same time.
+Also, countries typically decreased in inequality before increasing in environmental impact (Fig \ref{fig:cube_ensamble}A).
+Similarly, countries tended to see reductions in inequality before prosperity rose (Fig \ref{fig:cube_ensamble}B).
+Importantly, many middle-income countries had levels of inequality similar to high-income countries, but at lower levels of environmental impact and prosperity (Fig \ref{fig:cube_ensamble}B, C).
+Overall, our findings suggest that higher prosperity usually comes at the expense of environmental integrity.
+National trajectories through the three-dimensional trilemma space  also revealed that some countries struggled to escape areas characterized by high inequality, which may represent an “inequality trap” subset of the larger sustainability trilemma.
 
-The potential mechanisms of tradeoffs and synergies remain an important research frontier.
-Notably, tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC).
+It is important to note that this study was a historical, not predictive, analysis, and that past trajectories do not necessarily determine plausible future pathways.
+The potential mechanisms behind these tradeoffs and synergies between prosperity, inequality, and environmental impact constitute an important research frontier in sustainability science and development studies.
+The patterns and typologies identified in this paper could provide insights or directions for future investigations.
+Nonetheless, some mechanisms have already been hypothesized for pairwise relationships among the indicators in the often disparate literatures of economics, sociology, and environmental science.
+In what follows, we review several of the most influential hypotheses in the context of our findings.
+
+Tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC).
 In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955].
-Since this relationship was first posited, there has been conflicting evidence, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020].
-This has made the identification of mechanism difficult.
-Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or government regulations  [@Pikett:2014].
-Our analysis found no evidence for a KC-like relationship, instead finding that increasing inequality tended to precede growing prosperity.
+The evidence for this hypothesis has been conflicting, with some country-level studies finding strong correlations and others detecting no clear relationships [@Chancel:2020].
+This has made the identification of mechanisms difficult.
+Indeed, sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or concerted government actions  [@Pikett:2014].
+Our analysis found no evidence for a KC-like relationship in our study period, which was a time of rapid economic growth in many parts of the world.
 
-In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase their environmental impact with growing prosperity but after reaching an inflection point of affluence, the impact of an increasingly rich society decline [@barbier1997economic].
-As with the original KC, empirical studies of a wide range of environmental variables have come to conflicting conclusions, with some finding international evidence for this inverted-U-shaped relationship and others not [@dasgupta2002confronting; @stern2004rise].
-In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC, i.e.
-an inflection point in the prosperity-environment plane (Fig \ref{fig:cube_ensamble}).
-Various mechanisms have been proposed to explain this relationship or its absence, including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325].
-But what is clear from diverse empirical studies is that the connections between prosperity and environmental integrity are mediated or confounded by other, largely social, factors.
+In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence[@barbier1997economic].
+As with the original KC, empirical studies of a wide range of variables have come to conflicting conclusions.
+While some identified international evidence for this inverted-U-shaped relationship, others did not [@dasgupta2002confronting; @stern2004rise; @pothen:2019].
+In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we also did not detect an EKC-type trend: i.e., an inflection point in the prosperity-environmental impact plane (Fig \ref{fig:cube_ensamble}).
+Various mechanisms have been proposed to explain this relationship (or its absence), including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325].
+But what is clear from diverse empirical studies is that the connections between prosperity and environmental impact are mediated or confounded by other, largely social and institutional, factors.
 
-Other mechanisms and pathways that do not posit linear or quadratic relationships have been identified as driving interactions between the economic, social and environmental dimensions of sustainability.
+Notably, structural inequities in the global economy are creating or exacerbating a skewed distribution of development burdens.
+Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses [@MALER201325].
+For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013].
+Additionally, it has been found that international trade significantly influences global land use patterns, worsening natural resource exploitation and biodiversity loss [@Creutzig:2019].
+
+Within the trilemma space, there were international clusters that moved in distinctively different directions.
+These divergences further highlight the inadequacy of reductionistic, especially mono-causal, explanations (such as those found in many KC and EKC studies) and the need to incorporate historical and geographic factors as well.
+For instance, at the end of the study period, Ukraine had (relatively) low levels of inequality and environmental impact.
+However, the country’s  trajectory through the trilemma space was turbulent: over the study period, indicators rose and fell in dramatic downswings due to market shocks and corruption, endemic across many post-Soviet countries [@Duvanova:2014], as well as the social and economic dislocations caused by violent conflict.
+It is noteworthy that it was only in 2013  when Ukraine’s ecological footprint fell below its biocapacity (Fig.
+\ref{fig:S4}).
+Countries that performed relatively well across all three dimensions tended to be advanced Western economies, particularly those along the North Atlantic rim (e.g., Canada, Sweden).
+These countries had a much higher ratio of space and resources to people, as reflected in per-capita biocapacity.
+
+Interactions between the structure of the global economy, geographic conditions, and historical contingencies can help explain international differences with respect to the sustainability trilemma.
+This is perhaps best illustrated by the country that underwent the greatest magnitude of change in all three dimensions during the study period.
+In China, the rapid rise of prosperity, inequality, and environmental impact were significantly driven by the country’s integration into international commercial networks (Fig.
+\ref{fig:S4}).
+The rapid growth of incomes was underpinned by a resource-intensive export sector while rising income inequality was due in large part to geographic segmentation: urban coastal areas were the affluent export hubs while more rural inland areas supplied raw materials, intermediate goods, and cheap labour [@xie2014income; @qi2013interpreting].
+These changes, and the subsequent surges in all three dimensions, were set in motion after the tumultuous historical events of the 1960s and 1970s, which prompted the national leadership to undertake extensive economic reforms.
+
+Additionally, complex interactions among the three dimensions of the trilemma - and their implications for development - have also been recognized in interdisciplinary studies (none of which posit simple linear or quadratic relationships between the economic, social, and environmental dimensions of sustainability).
 For instance, Weinzettel et al.
 [@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
-Similarly, population, environmental carrying capacity, and pollution exhibit mutually-reinforcing feedbacks across countries [@LaTorr:2019].
-These types of feedbacks can create and maintain "red loop" or "green loop" dynamics at the national level, entrapping some states in long-term conditions of environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018; @Hamann:2015].
+Similarly, population, environmental carrying capacity, and pollution exhibited mutually-reinforcing feedbacks across countries [@LaTorr:2019].
+These types of feedbacks can create and maintain "red loop" or "green loop" dynamics between poverty and the environment, entrapping some countries in long-term conditions of high poverty and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018; @Hamann:2015].
 
-Additionally, structural inequities in the global economy are creating, or at least exacerbating, a skewed international distribution of sustainability-related development burdens.
-Environmental impacts can be "outsourced", as is common with affluent countries that have shifted resource-intensive production to developing countries through global supply chains [@Lambin:2011] - an important factor not captured in most EKC analyses [@MALER201325].
-For instance, the US and a number of European countries have outsourced a significant portion of their carbon emissions, land use, and environmental impacts to lower-income economies through trade [@Nijdam:2005; @Peters:2006; @Weinzett:2013].
-Additionally, it has been shown that international trade has significantly influenced global land use patterns, worsening natural resource exploitation and biodiversity loss[@Creutzig:2019].
-This "telecoupling" of unsustainable resource use between far away places may be a mechanism that can obfuscate real trajectories in the trilemma space, undermining the ability to reach all three goals at the global level by entrapping certain countries in undesirable regions.
+Although our analysis identified important patterns and drivers of international development (and corresponding lacunae in discipline-specific literature), it still has several limitations.
+First, the offshoring of environmental degradation discussed above is not completely captured in standard ecological footprint measurements [@VandenBergh:2012].
+A consumption-based approach to measuring environmental impacts – as opposed to the more mainstream production-based approach – may help internalize at least some of these externalities (e.g., many resource-intensive goods produced in China and Vietnam are consumed in the US and Europe, but the increases in ecological footprint thus generated are attributed to the former two).
+However, such an approach is constrained by the lack of relevant data, with  existing literature tending to be more narrowly focused on specific commodities and on bilateral or regional trade.
+If these outsourced impacts were incorporated into the appropriate indicators, our findings would likely show that affluent countries have been even less successful in overcoming the sustainability trilemma.
+Addressing this issue has important implications for the international distribution of responsibilities with respect to global commons challenges such as  mitigating climate change, biodiversity loss, and pandemic risks - all of which disproportionately affect lower-income countries [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].
 
-While no country was found to have eluded the sustainability trilemma, there were clusters that moved in distinctively different directions with respect to tradeoffs and synergies between the three goals.
-These dynamics further highlight the inadequacy of reductionistic, often mono-causal, explanations - such as those found in many KC and EKC studies - and the need to incorporate historical, geographic, and institutional factors.
-For instance, at the end of the study period, Ukraine had (relatively) low levels of inequality and environmental impact.
-However, the country’s  trajectory through the trilemma space was far from smooth: over the study period, indicators  rose and fell in dramatic downswings due to market shocks and corruption (endemic across post-Soviet countries throughout the 1990s) as well as  dislocations caused by war.
-In fact, it was only after the Russo-Ukrainian conflict began in 2013 that Ukraine’s ecological footprint fell below its biocapacity (Fig.
-\ref{fig:S4}).
-
-The clustering also shows the importance of geographic factors, including the natural resource endowment factor mentioned earlier.
-Countries that performed relatively well along all three dimensions tended to be advanced Western economies, particularly along the North Atlantic rim (e.g., Canada, Sweden).
-These countries had a much higher ratio of space and resources to people, as reflected in per-capita biocapacity.
-Finally, the structure of the global economy is also a powerful factor in explaining international differences in the trilemma space.
-In China, the rapid rise of both prosperity, inequality, and environmental impact were significantly driven by its integration into international commercial networks (Fig.
-\ref{fig:S4}).
-Rapid economic growth was driven by a resource-intensive export sector while rising income inequality was due in large part to geographic differences: urban coastal areas were the affluent export hubs while more rural inland areas supplied mostly raw materials, intermediate goods, and cheap labour [@xie2014income; @qi2013interpreting].
-
-Although our analysis identified important patterns and drivers of international development (as well as lacunae in the existing literature), it still faced practical limitations.
-First, the offshoring of environmental degradation discussed above is not completely captured in standard ecological footprint measurements [@VandenBergh:2012].
-A "bottom-up" or consumption-based approach to measuring environmental impacts – as opposed to the more mainstream "top-down" or production-based approach – may help internalize these externalities (e.g., resource-intensive goods produced in China are consumed in the US, but the environmental footprint of that production is attributed to the former).
-But such an approach is constrained by the lack of relevant data (the existing literature on this topic has more narrowly focused on specific commodities and on bilateral or regional trade).
-If these outsourced environmental impacts were incorporated into the appropriate indicators, our findings would likely show that affluent countries have been even less successful at reducing their ecological footprints.
-This phenomenon has important implications for the international distribution of ethical burdens, especially since climate change impacts, pandemic risks, and other global environmental problems will disproportionally affect lower-income countries, which tend to have lower adaptive capacity [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].
-
-A representative indicator-based approach to assessing sustainability at the national level has been used in other studies, yielding instructive lessons for research and policy [@Holden:2014; @Moran:2008; @Pulsell:2015].
-However, we recognize that there are other valid ways to conceptualize sustainability.
-Modern studies on the relationship between ecological limits, human population, and the economy have a long pedigree [@Ehrlich:1971; @Meadows:1972].
-More recently, an influential work by Raworth [-@RaworthKate:2017] postulated 11 "social thresholds" and 7 planetary boundaries to delimit sustainability.
-This and similar frameworks imply equal weight among chosen thresholds, which mix more subjective indicators (e.g., life satisfaction and democratic quality) with objective measures of wellbeing (e.g., nutrition and life expectancy).
-Such diverse classification schemes are valuable but may also be overdetermined: it is conceivable that a different and even longer list of thresholds could be advanced to form an equally relevant sustainability framework.
-In addition, there have been calls for global frameworks that move beyond the artificial separation of social and ecological goals towards more integrated social-ecological targets for sustainable development, to better account for the foundational role played by nature [@reyers2020global].
-While acknowledging that our three canonical dimensions framework is imperfect (or perhaps just coarse-grained), we believe its conceptual parsimony nonetheless reflects long-standing definitions of sustainability.
+We also recognize that there are other valid ways to conceptualize sustainability.
+For instance, studies on the relationship between ecological limits, human population, and the economy have a long pedigree [@Ehrlich:1971; @Meadows:1972].
+More recently, a popular  framework put forward by Raworth [-@RaworthKate:2017] postulated 11 "social thresholds" and 7 planetary boundaries to delimit sustainability.
+This framework and similar approaches imply equal weight among chosen thresholds, which mix subjective indicators (e.g., life satisfaction and democratic quality) with objective measures of wellbeing (e.g., nutrition and life expectancy).
+Such classification schemes are informative but may also be overdetermined: it is conceivable that a different and even longer list of thresholds could be advanced to form an equally relevant sustainability framework.
+In addition, there have been calls for analyses to better integrate social and ecological goals in order to account for the foundational role played by nature [@reyers2020global].
+While acknowledging that our “classic” three-pillars framework (sometimes termed the “triple bottom line”) is imperfect, we believe it nonetheless captures long-standing and important dimensions of sustainability and how countries have developed with respect to them.
 
@@
 
-Our analysis found that countries around the world face a trilemma when it comes to achieving economic prosperity and social equality while reducing environmental impact.
-While national strategies to navigate the trilemma space will be context- and path-dependent, our findings show that there are common typologies of how countries have historically evolved in the trilemma space.
-These typologies may help identify country groupings for which common development strategies address the challenges of  the sustainability trilemma.
-Nonetheless, it is important to recognize the differing geographic, historical, and institutional constraints facing countries, and that when it comes to the multi-dimensional challenges of sustainable development, there are no "one-size-fits-all" solutions [@Ostrom:2007; @ostrom2010moving; @perrings2007future].
+Our study found that countries around the world face a trilemma when it comes to achieving economic prosperity and social equality while reducing environmental impact - three goals that development scholarship and international agendas have defined as being fundamental to sustainability.
+It is important to recognize how institutional constraints, geographic differences, and historical contingencies and their interactions vary across countries.
+When it comes to such complex, multi-dimensional challenges, there are no "one-size-fits-all" solutions [@Ostrom:2007; @ostrom2010moving; @perrings2007future].
+We therefore argue that national approaches to overcoming the sustainability trilemma will need to be context-specific and must account for path-dependence.
+At the same time, our findings indicate that there are typologies of how countries have historically evolved with respect to prosperity, inequality, and the environment.
+These typologies can help identify common priorities and development strategies.
+This may be particularly relevant for Global South countries, which appear to face the steepest challenges with respect to the sustainability trilemma.
 
-Whether the patterns of synergies and tradeoffs identified in this study hold across scales (e.g., looking at finer subnational scales or other analytic units such as cities or corporations), remains an open question.
-Most importantly, the causal mechanisms explaining the synergies and tradeoffs between inequality, prosperity, and environmental impact are likely to be diverse, rather than mono-causal relationships.
-The foundational role of nature, which provides the basic life-support systems for humanity, cannot be understated when thinking about sustainable development goals.
-As our findings show, a myopic focus on prosperity is clearly misplaced - neither KC or EKC relationships were found for countries over the twenty-year study period.
+Whether the synergies and tradeoffs between prosperity, inequality, and environmental impact identified in this study hold across scales (e.g., at finer subnational scales or in other analytic units such as cities or corporations), remains an open question.
+Furthermore, these mechanisms  are likely to be complex rather than mono-causal [@Weinzett:2013; @Cumming:2018; @Hamann:2018; @LaTorr:2019].
+Additionally, the foundational role of nature, which provides the basic life-support systems for humanity, cannot be understated when thinking about sustainable development.
+As our findings strongly suggest, a myopic focus on ever-growing prosperity as the solution to inequality and environmental degradation is misplaced - neither KC or EKC relationships were found.
 Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts.
-An overemphasis on  economic prosperity can undermine progress towards the noble ambitions of leaving no country or person behind in the pursuit of global sustainability.
+One undeniable trend from recent history is that a singular focus on expanding income  can undermine progress towards the ambition of leaving no country or person behind in the pursuit of global sustainability [@arrow2004we; @fanning2022social].
 
-# Acknowledgements
-
-We would like to thank the conveners and other participants of the Beijer Young Scholars workshops (2016-2018) for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support.
-This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.
-
-**Author contributions:**  Conceptualization was a team effort, TW and JCR produced the first manuscript, TW, JCR and JQ analyized the data. All authors read and edit the paper.
+Overcoming the sustainability trilemma, for individual countries and the world as a whole, demands development that focuses as much on the distribution and impact of the economy as on its continued growth.
+This will likely require much greater investment in the social and natural capital that have been degraded, or allowed to atrophy, across the world in recent decades [@pretty2003social; @brondizio2009connectivity; @wackernagel1997perceptual; @Ehrlich:2012].
+Through this more inclusive approach, countries could still become more prosperous - just in a way that distributes those gains more fairly, which would help resolve the poverty and precarity that still bedevil so many, and with less impact on the biosphere, which is the ultimate guarantor of humanity’s prospective flourishing.
 

### 2023-07-04 paper/manuscript.Rmd
changes in response to reviewers

@@
 author: |
-    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^, Emilie Lindkvist^2^, Maike Hamann^6^, Jiangxiao Qiu^7^, 
+    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^,  Maike Hamann^6^, Emilie Lindkvist^2^, Jiangxiao Qiu^7^, 
     | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
     | \small
-    | \footnotesize ^1^The Natural Capital Project and Stanford Center on China's Economy and Institutions, Freeman Spogli 
-    | \footnotesize Institute for International Studies, Stanford University
+    | \footnotesize ^1^The Natural Capital Project and Institute for International Studies, Stanford University
     | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
@@
 #   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
-
 output:
@@
     toc: no
+  word_document:
 csl: /Users/juanrocha/Documents/styles/apa.csl
@@
 
-- Historical data reveals that no country in the world has reduced inequality and environmental footprint simultanously
-- A trilemma exist where countries need to trade-off progress on social and environmental goals
-- An inequality trap seems to exist particularly acute in Latinamerica
-- Countries trajectories on the trilemma space suggest that in the future social progress can be achieved without compromising the biosphere
+- Historical data reveals that no country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact.
+- A sustainability trilemma seems to exist where progress on either social goal comes at the expense of the environment.
+- An inequality trap seems to be particularly acute in Latin America.
+- Countries trajectories on the trilemma space suggest that future social progress can be achieved without compromising the biosphere.
 
@@
 
-For most of human history, material want, not sufficiency, was the norm for most people regardless of county [@Maddison:2007; @Harari:2014].
+For most of human history, material want, not sufficiency, was the norm for most people regardless of country [@Maddison:2007; @Harari:2014].
 Although global prosperity has increased in recent generations, progress has been uneven.
 If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium].
-And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]).
-Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and understandably aspires for higher and more secure living standards [@TheWor:2018].
-Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an increase rise of global poverty unprecedented since the end of World War II and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].
+And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]).
+Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and understandably aspire for higher and more secure living standards [@TheWor:2018].
+Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an increase of global poverty unprecedented since the end of World War II and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].
 
-At the same time, while average incomes for certain countries in what was once called the "Third World" began to  converge with those of the advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
+At the same time, while average incomes for certain countries in what was once called the "Third World" began to  converge with those of the advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both, developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
 The widespread feelings of injustice that the increasingly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009].
 Over the past decade, inequality has once again become an increasingly urgent issue for policymakers as well as the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020].
-These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks[@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017].
-Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened income and wealth inequalities within many countries, as well as widening the development trajectories between high- and low-income economies [@WorldBank:2022].
+These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks [@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017].
+Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened income and wealth inequalities within many countries, as well as widened the development trajectories between high- and low-income economies [@WorldBank:2022].
 
@@
 We explore trajectory typologies in the trilemma space (E-G).
-53 trajectories from 1995 to 2017 are depicted.
-The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (E), inequality and prosperity (F), and ecological footprint and prosperity (G).
+53 country trajectories from 1995 to 2017 are depicted.
+The trilemma is projected in each of the facets of the tridimensional cube: ecological footprint and prosperity (E), inequality and prosperity (F), and inequality and ecological footprint (G).
 Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
@@
 We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time.
-We further investigated whether these movements trace out typologies of countries within the trilemma space.
+We further investigated whether these movements trace out typologies of country trajectories within the trilemma space.
 To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.
 We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters).
-We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz].
+We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz].
 We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible).
@@
 The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
-Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values and then imputed them with a linear interpolation.
+Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $ fewer than 30\% of missing values and then imputed them with a linear interpolation.
 This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).
@@
 There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
-The inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters.
+The prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters.
 Despite this clustering, countries showed distinctive profiles in terms of their direction of movement (Fig \ref{fig:cube_ensamble} insets).
@@
 
-In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence[@barbier1997economic].
+In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence [@barbier1997economic].
 As with the original KC, empirical studies of a wide range of variables have come to conflicting conclusions.
@@
 For instance, Weinzettel et al.
-[@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
+[-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
 Similarly, population, environmental carrying capacity, and pollution exhibited mutually-reinforcing feedbacks across countries [@LaTorr:2019].

### 2023-07-04 revised_manuscript_230703.Rmd
changes in response to reviewers

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^,  Maike Hamann^6,7^, Emilie Lindkvist^2^, Jiangxiao Qiu^8^, 
    | \small Caroline Schill^2,9^, Alon Shepon^10^, Anne-Sophie Crépin^9^, and Carl Folke^2,9^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^Centre for Geography and Environmental Science, University of Exeter
    | \footnotesize ^8^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^9^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^10^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
  word_document:
csl: /Users/juanrocha/Documents/styles/apa.csl
bibliography: paper/references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

```{r data, include=FALSE}
library(tidyverse)
library(here)

## word counts
## Abstract 150
## Intro 697
## Method 957
## Results 621
## Discussion 1434
## Conclusion 267
## Total main 3976 + 150 abstract
## 
## Targets: 
# 
# - NatComms: 5000w | 150w abstract | 4 figs or tables | Methods end 3000w | Refs
# 70max
# 
# - WorldDev: | 4500-10k words including all (refs, captions, tables, etc) | 300w
# abstract
```

\doublespacing

**Author contributions:**  Conceptualization was a team effort, TW and JCR produced the first manuscript, TW, JCR and JQ analyzed the data. All authors contributed to the writing and editing of the paper.

**Declaration of interests:** None

**Correspondence:** juan.rocha@su.se | +46730747766

\pagebreak

# Highlights

- Historical data reveals that no country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact.
- A sustainability trilemma seems to exist where progress on either social goal comes at the expense of the environment.
- An inequality trap seems to be particularly acute in Latin America.
- Country trajectories in the trilemma space suggest that future social progress can be achieved without compromising the biosphere.

\pagebreak
# Acknowledgements

We would like to thank the conveners and other participants of the 2016-2018 Beijer Young Scholars workshops for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support. This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.

\pagebreak

# Global tradeoffs between prosperity, inequality, and the environment

\begin{abstract}

  A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact. But what if there is a "sustainability trilemma" that inhibits the simultaneous achievement of these three objectives? We analyzed synergies and tradeoffs between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017. Based on these widely-used measures of development, we found that no country overcame the hypothesized trilemma at the intersection of prosperity, equality, and environment. The trajectories of countries in the three-dimensional space formed by these key indicators revealed typologies and common development barriers. Over the study period, countries tended to decrease in income inequality before growing in prosperity and environmental impact. Some countries were trapped in a space of high inequality, while the instances of prosperous countries lowering inequality coincided with increasing environmental impact. The existence of clusters and widely divergent trajectories of certain countries within the trilemma space point to the need for explanatory frameworks that incorporate institutional, historical, and geographical drivers, as well as social-ecological feedbacks that strengthen their effects. Practically, our findings suggest that to overcome the sustainability trilemma, many countries - especially those that are already prosperous - should prioritise social redistribution and environmental stewardship before further economic growth.

\end{abstract}

# Introduction

Sustainability has often been characterised as having three essential and nested dimensions: economy, society, and the environment - what is sometimes referred to as the "triple bottom line". More specifically, a key development goal within this tripartite framework is to achieve high material living standards (prosperity) that are also well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008]. Programmatically, the United Nations’ Sustainable Development Goals explicitly recognize the important nexus formed by prosperity, equality, and the environment  [@Sachs:2015]. And in its report charting the path towards a sustainable recovery in the aftermath of COVID-19, the World Bank also identified the need to achieve "shared prosperity" by reducing inequality and accounting for the climate crisis and other environmental challenges [@WorldBank:2022]. Although each objective is desirable in its own right, whether prosperity, equality, and environmental integrity can be simultaneously achieved remains an empirically unanswered question. In other words, what if the achievement of the sustainability triple-bottom-line is impeded by the existence of a sustainability "trilemma"?

In the literature on international development, there have been many hypothesized "trilemmas" involving separately desirable but mutually incompatible objectives. For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination [@RodrikD:2011]. Using resources efficiently, fairly and with maximum freedom has been proposed as an unattainable triad [@Coyle:2011]. Similarly, structural tradeoffs between prosperity, equality, and the environment  have been investigated in different locations using various indicators. A statistical analysis found that economic inequality and carbon emissions were negatively correlated in low- and middle-income countries but positively correlated in high-income countries [@Grunewald2017]. In contrast, an earlier econometric study concluded that mitigating carbon emissions trades-off with both economic growth and social equity [@Ravallion_2000]. And focusing on three case studies in the United States and Southeast Asia, Halpern et al. (2013) found that achieving biodiversity objectives is often facilitated by the inequitable distribution of resources, complicating the achievement of "triple-bottom-line solutions" in conservation planning [@Halpern_2013]. Using an agent-based model of collective decision-making that assumes simple majority rule and takes for granted prosperity-environment conflicts, Kempf and Rossignol (2007) argued that widening inequality undermines public willingness to support environmental policies [@KEMPF_2007].

Despite earlier studies on potential sustainability trilemmas, few analyses have taken a dynamic global perspective. Many of the most critical development challenges in the Anthropocene, when human activities have become the primary drivers of planetary change, are transboundary and global in scale. Identifying the distribution of these challenges across countries is key to arriving at principles for collective action. At the same time, it is important to track how different countries have progressed or regressed with respect to key objectives over meaningful time-frames. In this way, country-specific as well as shared international obstacles and opportunities can be identified, helping to inform the design and implementation of policies.

In this study, we ask whether there is evidence – and rationale – that the three sustainability goals of prosperity, equality, and environmental integrity can be jointly achieved by any given country using an international, time-series dataset of representative development indicators. In other words, this study attempts to answer the questions: To what extent does such a sustainability trilemma exist and are there inherent tradeoffs between its constituent goals? If such a sustainability trilemma does exist, where does it occur and why does it occur there? By providing answers to these questions, we hope to contribute to the increasingly urgent discourse on how to achieve sustainable and inclusive prosperity in an age of unprecedented planetary change.

## Background

For most of human history, material want, not sufficiency, was the norm for the majority of people, regardless of country [@Maddison:2007; @Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]). Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018]. Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an unprecedented increase of global poverty and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

At the same time, while average incomes for certain countries in what was once called the "Third World" have converged with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].

The widespread feelings of injustice that highly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009]. Over the past decade, inequality has once again become an increasingly urgent issue for policymakers and the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020]. These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks [@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017]. Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened economic inequalities within many countries, as well as widened the development trajectories between high- and low-income economies [@WorldBank:2022].

Environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic]. Interdisciplinary research confirms that human well-being and economic development depend on the resilience of nature. There is also a scientific consensus that human activities have a wide and deleterious impact on the biosphere, which constitutes the habitative upper bound and common denominator for our species (notwithstanding fanciful visions of extraterrestrial expansion)  [@Díaz:2019; @folke2021our]. Anthropogenic drivers have become primarily responsible for global biodiversity loss, land degradation, and climate change, among other environmental impacts [@Díaz:2019]. Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023]. Global development in the 21st Century therefore urgently requires more sensible stewardship of the biosphere, from ecosystems to the climate [@Millenniu:2005; @IPCC:2021].

# Methods

To assess the extent, distribution, and evolution of the hypothesized sustainability trilemma at the global level, we examined the performance of countries over the past quarter-century using publicly available data for three representative and widely-used indicators of development (Table \ref{tab1}). We chose the country level because many of the most consequential policies, plans, and institutions affecting sustainability are made by national governments [@Bryan:2018], and also due to practical considerations of data availability. We operationalized the sustainability trilemma as the movement of countries over time in the three-dimensional space formed by the axes of prosperity, (in)equality, and environmental integrity (alternatively, impact)  (Fig \ref{fig:trilemma}). There is a development synergy if a country moves in desired directions towards more than one goal (e.g., reducing inequality and increasing prosperity). There is a development tradeoff if positive movement towards one goal is accompanied by movement in an undesirable direction for another. Therefore, synergies between two goals are movements along the (blue) diagonal (positively correlated), while tradeoffs are represented by the (red) axis, where the achievement of one hinders the achievement of another (negatively correlated). It is important to note that in this framing, the sign of the correlation changes depending on the term used. For instance, if one speaks of equality as opposed to inequality, or environmental integrity as opposed to environmental impact. In the remainder of this paper, we use inequality and environmental impact because this is how the original data was defined and structured. Therefore, across these two domains, increasing numbers mean less desirable conditions, and negative correlations translate to synergies between the goals of prosperity, equality, and environmental integrity  (Fig \ref{fig:trilemma}).

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{fig1_trilemma_space}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A). There could be synergies and trade-offs in achieving these three objectives (B-D). We explore trajectory typologies in the trilemma space (E-G). 59 country trajectories from 1995 to 2017 are depicted. The trilemma is projected in each of the facets of the tridimensional cube: ecological footprint and prosperity (E), inequality and prosperity (F), and inequality and ecological footprint (G). Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
\label{fig:trilemma}
\end{figure*}
```

To quantify prosperity, inequality and environmental impact, we identified a set of widely-used and publicly-available indicators (Table \ref{tab1}). For prosperity, we used per-capita gross national income (GNI) in US\$ values. GNI is related to gross domestic product (GDP) and the two are close in value for most countries. A key difference is that GNI accounts for earnings from foreign investments. The World Bank uses GNI per capita as its standard measure for the level of development (e.g., low-, medium- or high-income, data source: <http://data.worldbank.org>). The threshold for being "prosperous" was US\$12,746 per capita per year, which was the World Bank's definition for a high-income country in 2017, the concluding year of our study period [@TheWor:2018].

To measure inequality, we used the Gini coefficient for income. The Gini coefficient is a statistical measure of dispersion and has a long history of being used to measure the level of economic inequality within a country [@Pyatt:1976]. Hypothetically, a country with a Gini value of 0 is absolutely equal, while a value of 1 represents absolute inequality. Gini coefficients of income were taken from the World Income Inequality Database v. 3.4 (source: [https://www.wider.unu.edu/data](https://www.wider.unu.edu/data)). We used 0.30 as the threshold below which countries are identified as having low inequality, based on assessments from historical economic studies [@Piketty:2014].

We used per-capita ecological footprint to measure environmental impact. Ecological footprint is a holistic estimate of the biophysical resources used by human activity, quantified in terms of the land area required for the extraction of materials and energy and the absorption by the environment of pollution and waste [@Wackernagel:1998]. Its unit is the "global hectare", combining measurements of carbon emissions, built-up land, fishing grounds, croplands, forests, and grazing land into an index of environmental impact. In conjunction with measurements of biocapacity – or the extent to which the environment can provide, absorb, and regenerate – the ecological footprint determines whether a given country (or person or city or the world as a whole) is in an unsustainable condition of ecological "deficit". Although it may be too aggregative of the diverse facets of anthropogenic environmental change, the ecological footprint remains the most widely-used and data-rich approximation of overall human impact on the biosphere in a single metric, while also accounting for the effects of trade on resource consumption by country. Ecological footprint and biocapacity data were taken from the Global Footprint Network (source: [https://www.footprintnetwork.org/resources/data/](https://www.footprintnetwork.org/resources/data/)). We used 1.68 global hectares per capita as the threshold for environmental impact; any value higher would exceed global biocapacity (i.e., leading to ecological deficit). In other words, it is the limit at which current generations' needs can be satisfied without compromising the ability of future generations to satisfy their needs.

```{r tab1, echo = FALSE, include = TRUE, results = "asis"}
dat <- tibble(
    Dimension = c("Prosperity", "Equality", "Environment"),
    Indicator = c(
        "Per-capita gross national income in US$",
        "Gini coefficient for income",
        "Per-capita ecological footprint in global hectares"),
    N = c(232, 196, 211),
    Threshold = c(12746, 0.3, 1.68)
)

stargazer::stargazer(
    dat, type = "latex", 
    title = "Indicators used in the assessment of national prosperity, equality and environmental impact.",
    label = "tab1", summary = FALSE, header = FALSE)

# knitr::kable(dat, caption = "Indicators used in the assessment of national prosperity, equality and environmental impact.", label = "tab1")
```

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of country trajectories within the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time. We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

While we were able to assess 140 countries "statically" (i.e., ignoring temporal dynamics), an analysis of the missing values showed that analyzing changes over time in the trilemma space was only possible for 59 countries. The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected the time-series from 1995-2017 for the 59 countries with fewer than 30\% of missing values and then imputed the missing values with linear interpolation. This reduced dataset enables the study of the temporal dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

# Results

Our analysis of the three representative indicators and their respective thresholds found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact. This strongly suggests that globally, there is a sustainability trilemma. At the country level, there were strong tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction). Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden and New Zealand) exhibited low levels of inequality in the past but then high variation and a trend towards greater inequality in more recent years.

While these globally-aggregated patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies. To further investigate individual trajectories, we calculated the displacement within the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane. These two angles and the total distance "travelled" in the three-dimensional space help to define typologies in country trajectories.

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{trilema_ensamble}
\caption{\textbf{Countries' positions and trajectories within the trilemma space} 59 country trajectories from 1995 to 2017 are depicted. Colours and contours represent areas of similar values and enclose attractors within peaks. The insets of each graph show the direction of movement for countries during the study period.}
\label{fig:cube_ensamble}
\end{figure*}
```

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). Building on insights from feedback dynamics described in other fields of development studies and sustainability science, multiple stable states within the trilemma space are plausible if feedbacks between development drivers are present. Statistically, this translates into multi-modal distributions and so-called "attractors". The prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Countries also showed distinctive profiles in direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three trilemma indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

The trajectories of countries in the trilemma space from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A). First, there was a group of countries, mainly from Latin America, that were "trapped" in a region of high inequality. They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green). The second group showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow). The third group was composed of countries with relatively high prosperity and ecological footprint; their movements in the trilemma space were relatively less erratic, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).

If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma indicators, we are able to include more countries (N=140). This static analysis yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the dynamic analysis. The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies and had high prosperity, inequality, and environmental impact. The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "developing world" or "Global South". Here, countries all tended to have lower prosperity (e.g., low- or middle-income status), but among them exhibited large variations in inequality and environmental impact.

Decreasing inequality and decreasing environmental impact tended to pull countries in opposite directions (Fig \ref{fig:pca}), providing strong evidence for the difficulty of achieving both objectives at the same time. In Fig \ref{fig:pca}A the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences. While environmental footprint (EF) and gross national income (GNI) point in similar directions, meaning countries tend to increase both at the same time, our inequality proxy (Gini) points in the opposite direction, confirming our previous speculation: countries that reduce inequality do so by increasing their environmental footprint and prosperity. In Fig \ref{fig:pca}B the ordination is performed in a larger but static sample (140 countries), and we find a similar pattern although not completely in opposite directions. In fact, in our trajectories analysis we found that  countries that reduced inequality subsequently increased their environmental impact (Fig \ref{fig:cube_ensamble}A). Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B). However, some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact - albeit at lower levels of prosperity (Fig \ref{fig:cube_ensamble}B, C). Overall, our findings are in line with the conventional wisdom that higher prosperity usually comes at environmental expense. The low frequency with which countries were able to meet even two thresholds - none in the case of the prosperity-environment relationship - is notable. National trajectories through the three-dimensional trilemma space also revealed that some countries struggled to escape areas characterised by high inequality, which may represent an "inequality trap" subtype of the more general sustainability trilemma.

```{=tex}
\begin{figure*}[htp]
\centering
\includegraphics[width = 6.5in, height = 5in]{pca_maps}
\caption{\textbf{Country typologies on the trilemma space} Principal component analysis and a senstivity analysis on clustering routines are used to identify country typologies. Three clusters are identified when looking at the historical trajectory of countries (N=59) from 1995-2017 (A). Two clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness. It complements the dynamic analysis with one summary statistic over time for 140 countries (B). For both A and B, red arrows are vectors fitted to the data that indicate the directions along which countries and clusters differ from each other. Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
\label{fig:pca}
\end{figure*}
```

# Discussion

In this study, we conducted a retrospective, dynamic analysis, and it is important to note that past trajectories do not necessarily determine future pathways. The potential mechanisms of tradeoffs and synergies between prosperity, inequality, and environmental impact constitute an important research frontier in sustainability science and development studies. In the often disparate literatures of economic, social, and environmental science, no theory has so far been advanced to account for the country patterns and typologies identified here. Certain mechanisms have been hypothesized for pairwise relationships among the development indicators used in this study. However, only by considering all three dimensions together, and the different ways in which countries have evolved with respect to them, can we derive a more comprehensive picture of the sustainability challenges for international development. Before positing trilateral explanations for the existence and distribution of the sustainability trilemma, we will first review several of the most influential dilemma hypotheses and discuss why they may be at odds with our findings.

Tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC). In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955]. The evidence for this hypothesis has been contested [@Chancel:2020]. Sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or even concerted government actions [@Pikett:2014]. Our analysis of per-capita income and income distribution found no evidence for a KC-like relationship from 1995 to 2017, which was a period of rapid economic growth in many parts of the world (see SI).

In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) is a complementary hypothesis: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence [@barbier1997economic; @grossman1995economic]. As with the original KC, empirical studies of a wide range of variables have produced mixed findings. Some studies have identified evidence for this inverted-U-shaped relationship. For instance, EKC relationships were often found for local air pollutants but not for river pollution, deforestation, or global greenhouse gas emissions [@stern2004rise; @stern1996economic ]. In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC-type trend: i.e., an inflection point in the prosperity-environmental impact plane (Fig \ref{fig:cube_ensamble}, Fig \ref{fig:S5}). Various mechanisms have been proposed to explain this relationship (or its absence), including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325].

What is clear from these earlier empirical studies of pairwise relationships between prosperity, inequality, and the environment is that whatever tradeoffs and synergies that were identified are mediated by exogenous factors [@dasgupta2002confronting; @pothen:2019]. Indeed, in his in-depth survey of relevant studies, Chancel (2020) concludes that what empirical evidence has been found to support (E)KC-type relationships are actually artefacts particular to post-World War II datasets or transient historical trends with limited applicability to today’s inequality dynamics [@Chancel:2020]. Within the trilemma space, this need for more sophisticated explanatory frameworks was underscored by the existence of international clusters that moved in distinctively different directions. These dynamics highlight the inadequacy of reductionist explanations that do not incorporate institutional, historical and geographic factors and the social-ecological feedbacks that reinforce their effects.

A key institutional factor driving divergent international development outcomes over our study period was the structure of global commerce. In particular, it has become evident that globalization created or exacerbated the (already) skewed distribution of environmental impacts worldwide. Notably, affluent countries "outsource" high-throughput,- land-intensive, and environmentally-polluting production to developing countries through international supply chains [@Nijdam:2005; @Peters:2006; @Lambin:2011; @Weinzett:2013] - an important factor not captured in most EKC analyses [@MALER201325]. Additionally, it has been found that international trade influences global land use patterns, worsening natural resource exploitation and biodiversity loss [@Creutzig:2019]. Our study period of 1995-2017 coincided with an intensive period of globalization, stimulated by the end of the Cold War and driven by the integration of large, emerging economies into international commercial and financial networks.

With respect to historical factors, Ukraine provides a potentially instructive example. At the end of the study period, Ukraine had (relatively) low levels of inequality and environmental impact. However, the country’s  trajectory through the trilemma space was turbulent: over the study period, indicators rose and fell in dramatic downswings due to market shocks and corruption, which has been endemic in many post-Soviet countries [@Duvanova:2014], as well as the social and economic dislocations caused by violent conflict. It is noteworthy that it was only in 2013 that Ukraine’s ecological footprint fell below its biocapacity, accompanied, or more likely driven, by a precipitous decline in per-capita income  (Fig. \ref{fig:S4}). The importance of geographic factors is underscored by the fact that countries that performed relatively well across all three dimensions tended to be advanced Western economies, particularly those that had a much higher ratio of space and resources to people. This is reflected in these countries’ per-capita biocapacity, providing greater potential for the increased ecological footprint that is associated with economic growth.

The effects of globalization, historical contingency, and geographical constraints are perhaps best illustrated by the country that underwent the greatest magnitude of change in all three dimensions during the study period. In China, the rapid rise of prosperity, inequality, and environmental impact were significantly driven by the country’s integration into international circuits of trade and finance, especially after accession to the World Trade Organization in 2001 (Fig. \ref{fig:S4}). The rapid growth of incomes was underpinned by a resource-intensive export sector while rising income inequality was due in large part to geographic segmentation: urban coastal areas were the affluent export hubs while more rural inland areas supplied raw materials, intermediate goods, and cheap labour [@xie2014income; @qi2013interpreting]. These changes, and the subsequently dramatic changes in all three sustainability dimensions, were set in motion after the tumultuous historical events of the 1960s and 1970s, which prompted the national leadership to undertake extensive economic reforms.

Interdisciplinary studies have begun to highlight the role of social-ecological feedbacks in reinforcing the effects of historical contingency (generating hysteresis or path dependency) and geographic constraints. For instance, Weinzettel et al. [-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions. Similarly, population, environmental carrying capacity, and pollution exhibited mutually-reinforcing feedbacks across countries [@LaTorr:2019]. This helps to explain the relatively good performance of countries such as Canada, Sweden, and Australia in our analysis. These types of feedbacks can create and maintain "red loop" or "green loop" dynamics between poverty and the environment, entrapping some countries in long-term conditions of high poverty and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018].

Although this study identified important patterns of international development and posited their potential, underlying drivers, it still faced several limitations. First, the offshoring of environmental degradation through global supply chains discussed above is not fully captured in standard ecological footprint measurements [@VandenBergh:2012]. A consumption-based approach to measuring environmental impacts – as opposed to the more mainstream production-based approach – may help internalize at least some of these externalities (e.g., many resource-intensive goods produced in Asia are consumed in the US and Europe, but the consequent increases in ecological footprint are attributed to the former). However, such an approach is currently constrained by a lack of necessary data, with existing studies tending to focus more narrowly on subsets of commodities and on bilateral or regional trade. If these outsourced impacts were incorporated into the appropriate indicators, our findings would likely show that affluent countries have performed much more poorly with respect to the sustainability trilemma. Addressing this issue has important implications for the international distribution of responsibilities with respect to global commons challenges such as mitigating climate change, biodiversity loss, and pandemic risks - all of which disproportionately affect lower-income countries [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].

Whether the synergies and tradeoffs between prosperity, inequality, and the environmental impact identified in this study hold across scales (e.g., at subnational scales or in other analytic units such as cities or corporations), remains an open question. Furthermore, the feedback mechanisms for interactions among the three dimensions at different scales are also likely to be complex rather than mono-causal [@Weinzett:2013; @Cumming:2018; @Hamann:2018; @LaTorr:2019]. If there is a unifying principle or concern among countries and across scales, it may be the foundational role of nature, which provides the basic life-support systems for humanity and cannot be understated when thinking about sustainable development. As our findings strongly suggest, a myopic focus on ever-growing prosperity as the solution to inequality and degradation is misplaced - neither KC or EKC relationships were found in our analysis (SI). Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts. One undeniable trend from our analysis of recent world development is that a singular focus on economic growth can undermine progress towards social equality and, most especially, environmental integrity [@arrow2004we; @fanning2022social].

Finally, we recognize that there are other valid ways to conceptualize sustainable development. For instance, although they did not account for distributional issues, studies on the relationship between ecological limits, human population, and the economy have a long pedigree [@Ehrlich:1971; @Meadows:1972]. More recently, a popular framework put forward by Raworth [-@RaworthKate:2017] postulated 11 "social thresholds" and 7 planetary boundaries to delimit sustainability. This framework and similar approaches imply equal weight among chosen thresholds, which may mix subjective indicators (e.g., life satisfaction and democratic quality) with objective measures of wellbeing (e.g., nutrition and life expectancy). Such classification schemes are informative but may also be overdetermined: it is conceivable that a different and equally justifiable list of thresholds could be proposed. In addition, there have been calls for analyses to better integrate social and ecological goals in order to account for the foundational role played by nature [@reyers2020global]. While acknowledging that our "classic" triple bottom line framework is imperfect, we believe it nonetheless captures long-standing and important dimensions of sustainability and how countries have progressed or regressed within the developmental space formed by them.

# Conclusions

Our study found that countries around the world face a trilemma when it comes to improving economic prosperity and social equality while reducing environmental impact - three goals that the global development agenda has identified as being fundamental to sustainability. It is important to recognize how globalization, historical contingencies, and geographic constraints d - along with the feedbacks that strengthen their effects -  vary across countries and drive their development within the trilemma space. When it comes to such complex development challenges in an era of planetary change, there are no "one-size-fits-all" solutions [@Ostrom:2007; @ostrom2010moving; @perrings2007future]. We therefore argue that national approaches to overcoming the sustainability trilemma will need to be context-specific and must account for path-dependence. At the same time, our findings indicate that there are typologies of how countries have historically evolved with respect to prosperity, inequality, and the environment. These typologies can help identify common priorities and suggest development strategies. This may be particularly relevant for Global South countries, which appear to face the steepest challenges with respect to the sustainability trilemma.

Overcoming the sustainability trilemma, for individual countries and the world as a whole, demands development that focuses as much on the distribution and impact of the economy as on the economy’s continued growth. This will likely require much greater investment in the social and natural capital that have been degraded, or allowed to atrophy, across the world in recent decades, from advanced Western economies to the large "emerging markets" to the least developed states of the Global South - where healthy communities and resilient environments will be critical for their future in the Anthropocene [@pretty2003social; @brondizio2009connectivity; @wackernagel1997perceptual; @Ehrlich:2012]. Through this more inclusive approach, countries could still become more prosperous. But that prosperity should be achieved in ways that distributes gains more fairly, which would help resolve the poverty and precarity that still bedevil so many, and with less impact on the biosphere, which remains the ultimate foundation for humanity’s prospective flourishing.

# References {.unnumbered}

::: {#refs}
:::

\pagebreak

# Supplementary Material {#sec:SM}

## List of countries

**Dynamic analysis (N = 59):**

```{r results="asis", echo=FALSE, message=FALSE, warning=FALSE}

load("data/cleaned_trilemma-20230614.RData")
load("data/trilemma_static.Rda")
countries <- readr::read_csv(
    file = "~/Documents/Projects/DATA/WorldBank/SDG_csv/SDGCountry.csv") |> 
    janitor::clean_names()

df_dat <- df_dat |> 
    rename(EF = EFconsPerCap, GNI = gni2, Gini = gini)

c <- df_dat |> 
    left_join(countries, by = c("iso3" = "country_code")) |> 
    pull(short_name) |> unique()

cat(c, sep = ", ")
```

**Static analysis (N = 140):**

```{r results="asis", echo=FALSE, message=FALSE, warning=FALSE}
c <- df_all2 |> 
    left_join(countries) |> 
    pull(short_name) |> unique()

cat(c, sep = ", ")
```

```{=tex}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{prosperity_missing_values}
\caption{\textbf{Data quality for prosperity} Different prosperity metrics are compared with respect to their coverage in space (countries) and time (A). Gross national income (GNI) is the metric with higher completion rate (B), with an arbitrary threshold of 30\% missing values, the analysis could include over 100 coutries (C).}
\label{fig:SM_gni_missing}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{inq_missing_values}
\caption{\textbf{Data quality for inequality} Mean Gini income is the most complete statistic of inequality with respect to their coverage in space (countries) and time (A). Alternative metrics such as the Palma index overcome some of the limitations of Gini but have disproportionally more missing values (B). For Gini, the proportion of countries with missing values is above 50\% for any given year. An arbitrary treshold of 30\% threshold for includsion will leave countries like South Africa (country with highers inequality in the world) out of the analysis (D).}
\label{fig:SM_inq_missing}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 6in]{ecofp_missing_values}
\caption{\textbf{Data quality for ecological footprint} Coverage of the ecological footprint in space (countries) and time (A) shows that data is relatively complete after mid 1990s (B) An arbitrary threshold of 10 missing values would leave some of the small islands states out of the analysis (C).}
\label{fig:SM_ecofp_missing}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{FigS4}
\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the "trilemma space" formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
\label{fig:S4}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3.5in]{sm_trajectories}
\caption{\textbf{Trajectories per continent} 59 countries trajectories coloured by continent. Some countries are labelled but not all of them to avoid over plotting.}
\label{fig:S5}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3.5in]{first_diff}
\caption{\textbf{First difference trajectories} 59 countries trajectories as first difference per variable.}
\label{fig:S6}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 5in]{sm_kuznets}
\caption{\textbf{Kuznets hypotheses and our data}  The top panel shows the Kuznets curve (KC) hypothesis (A), the environmental (EKC) counterpart (B), and the logical consequence of a linear relationship if both KC and EKC are true (C). The central panel (D-F) shows a non-linear loess fit on our trajectories data (59 countries, 22 year period), and the bottom panel (G-I) the same fit for the static analysis (140 countries). E,F,H,I deviate completely from the expected hypotheses. D and G show a mild curvature (the coefficient is significant but is 10 orders of magnitude smaller than the linear counterpart, table \ref{tab:1}), and the regions of the curve one can trust less are the extremes that make up the curvature due to the expanding errors. In fact, the far left side of panel D are the first three years of the time series for Spain, Portugal and Italy. The data does not conform to the KC or EKC hypotheses.}
\label{fig:EKC}
\end{figure*}
```

```{r echo=FALSE, results="asis", message=FALSE, warning=FALSE}
#linear regression:
fit <- glm(EF  ~ GNI * Gini + I(GNI^2) + I(Gini^2), 
           family = "gaussian", 
           data = df_dat)

stargazer::stargazer(
    fit, header = FALSE,
    title = "Regression of the trilemma space",
    label = "tab:1")
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 3.5in]{sm_cube}
\caption{\textbf{3D cube of the trilemma space} The surface from regression in Table \ref{tab:1} is shown in gray, trajectory points are show in the three-dimensional space with colors proportional to the environmental footprint axes. The best fit differs from quadratic forms on the KC and EKC axes.}
\label{fig:sm_cube}
\end{figure*}
```

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 4in, height = 4in]{correlogram}
\caption{\textbf{Correlogram} Correlations between the dimensions of the trilemma space}
\label{fig:cors}
\end{figure*}
```

### 2023-07-04 submit.R

@@ -1,0 +1,12 @@
+library(pdftools)
+
+## Routine for first submission
+pdf_subset("paper/manuscript.pdf",pages = 1:3, output = "paper/title_page.pdf")
+pdf_subset("paper/manuscript.pdf", pages = 4:25, output = "paper/main_manuscript.pdf")
+pdf_subset("paper/manuscript.pdf", pages = 1, output = "paper/conflict_statement.pdf")
+
+
+## Routine for second sumbission
+pdf_subset("revised_manuscript_230703.pdf", pages = 4:32, output = "revised_main_manuscript.pdf")
+pdf_subset("response to reviewers.pdf", pages = 1, output = "letter_editor.pdf")
+pdf_subset("response to reviewers.pdf", pages = 2:7, output = "response_reviewers.pdf" )

### 2023-11-20 revised_manuscript_231115.Rmd
uploading replication datasets

--- 2023-07-04 revised_manuscript_230703.Rmd
@@
 ---
-title: "Global tradeoffs between prosperity, inequality, and the environment"
+title: "Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment"
 author: |
-    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^,  Maike Hamann^6,7^, Emilie Lindkvist^2^, Jiangxiao Qiu^8^, 
-    | \small Caroline Schill^2,9^, Alon Shepon^10^, Anne-Sophie Crépin^9^, and Carl Folke^2,9^
+    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,4§^, Kevin Berry^5^, Tomas Chaigneau^6^,  Maike Hamann^7,8^, Emilie Lindkvist^2^, Jiangxiao Qiu^9^, 
+    | \small Caroline Schill^2,4,10^, Alon Shepon^11^, Anne-Sophie Crépin^10^, and Carl Folke^2,4,10^
     | \small
-    | \footnotesize ^1^The Natural Capital Project and Institute for International Studies, Stanford University
+    | \footnotesize ^1^The Natural Capital Project, Doerr School of Sustainability, Stanford University
     | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
     | \footnotesize ^3^Future Earth
-    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
-    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
-    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
-    | \footnotesize ^7^Centre for Geography and Environmental Science, University of Exeter
-    | \footnotesize ^8^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
-    | \footnotesize ^9^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
-    | \footnotesize ^10^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
+    | \footnotesize ^4^ The Anthropocene Laboratory, The Royal Swedish Academy of Sciences
+    | \footnotesize ^5^College of Business and Public Policy, University of Alaska Anchorage
+    | \footnotesize ^6^Environment and Sustainability Institute, University of Exeter
+    | \footnotesize ^7^Centre for Sustainability Transitions, Stellenbosch University
+    | \footnotesize ^8^Centre for Geography and Environmental Science, University of Exeter
+    | \footnotesize ^9^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
+    | \footnotesize ^10^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
+    | \footnotesize ^11^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
     | \footnotesize ^§^These authors made equal contributions
@@
 
-- Historical data reveals that no country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact.
-- A sustainability trilemma seems to exist where progress on either social goal comes at the expense of the environment.
-- An inequality trap seems to be particularly acute in Latin America.
-- Country trajectories in the trilemma space suggest that future social progress can be achieved without compromising the biosphere.
+- No country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact between 1995-2017.
+- A sustainability “trilemma” as opposed a “triple bottom line” appears to exist for international development.
+- There is an “inequality trap” that appears to be particularly acute in Latin America.
+- Country trajectories suggest that reducing inequality can be achieved without compromising the biosphere.
+- The three dimensions need be considered together to identify international development challenges.
 
@@
 We would like to thank the conveners and other participants of the 2016-2018 Beijer Young Scholars workshops for their insightful comments and advice as well as the Beijer Institute of Ecological Economics, at the Royal Swedish Academy of Sciences, for its institutional support.
-This work was also supported by funding from Formas (grant number 2018-02318), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.
+This work was also supported by funding from Formas (grant number 2018-02318 and grant number 2020-00454), and the Swedish Research Council Dnr 2018-05862 and Dnr 2018-06139.
 
@@
 
-# Global tradeoffs between prosperity, inequality, and the environment
+# Triple bottom line or trilemma? Global tradeoffs between prosperity, inequality, and the environment
 
@@
 A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact.
-But what if there is a "sustainability trilemma" that inhibits the simultaneous achievement of these three objectives?
-We analyzed synergies and tradeoffs between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017.
-Based on these widely-used measures of development, we found that no country overcame the hypothesized trilemma at the intersection of prosperity, equality, and environment.
-The trajectories of countries in the three-dimensional space formed by these key indicators revealed typologies and common development barriers.
+This has often been called the “triple bottom line”.
+But instead, what if there is a " trilemma" that inhibits the simultaneous achievement of these three goals?
+We analyzed international patterns and trends in the relationships between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017.
+Based on these widely-used measures of development, we found that no country achieved the triple bottom line.
 Over the study period, countries tended to decrease in income inequality before growing in prosperity and environmental impact.
-Some countries were trapped in a space of high inequality, while the instances of prosperous countries lowering inequality coincided with increasing environmental impact.
-The existence of clusters and widely divergent trajectories of certain countries within the trilemma space point to the need for explanatory frameworks that incorporate institutional, historical, and geographical drivers, as well as social-ecological feedbacks that strengthen their effects.
+A cluster of countries, primarily from Latin America, appeared trapped in a regime of high inequality, while the instances of prosperous countries lowering inequality only coincided with increasing environmental impact.
+The existence of country clusters and widely-divergent national trajectories show that common pair-wise explanations such as Kuznets Curves do not fully capture important development dynamics, especially in a period of global environmental change.
+Instead, frameworks need to consider all three dimensions together, and especially the relationship between inequality and the environment.
+Additionally, the past development trajectories of countries, how they fit into the structure of the world economy, and the role of social-ecological feedbacks in strengthening clustering effects such as “inequality traps”.
 Practically, our findings suggest that to overcome the sustainability trilemma, many countries - especially those that are already prosperous - should prioritise social redistribution and environmental stewardship before further economic growth.
@@
 Sustainability has often been characterised as having three essential and nested dimensions: economy, society, and the environment - what is sometimes referred to as the "triple bottom line".
-More specifically, a key development goal within this tripartite framework is to achieve high material living standards (prosperity) that are also well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008].
+More specifically, an important development objective within this framework is to achieve high material living standards (prosperity) that are also well distributed (equality) and maintainable over the long run (environmental integrity) [@Ehrlich:2012; @Sachs:2015; @Tracey:2008].
 Programmatically, the United Nations’ Sustainable Development Goals explicitly recognize the important nexus formed by prosperity, equality, and the environment [@Sachs:2015].
-And in its report charting the path towards a sustainable recovery in the aftermath of COVID-19, the World Bank also identified the need to achieve "shared prosperity" by reducing inequality and accounting for the climate crisis and other environmental challenges [@WorldBank:2022].
-Although each objective is desirable in its own right, whether prosperity, equality, and environmental integrity can be simultaneously achieved remains an empirically unanswered question.
-In other words, what if the achievement of the sustainability triple-bottom-line is impeded by the existence of a sustainability "trilemma"?
+And in its report charting the path towards a sustainable recovery in the aftermath of COVID-19, the World Bank also identified the need to achieve "shared prosperity" by reducing inequality and addressing environmental crises [@WorldBank:2022].
+However, whether the three goals can be simultaneously achieved remains an open question – what if there is a "sustainability  trilemma" that prevents countries from reaching this development objective?
 
-In the literature on international development, there have been many hypothesized "trilemmas" involving separately desirable but mutually incompatible objectives.
-For instance, it has been argued that a country cannot achieve globalization, democracy, and sovereignty in a non-volatile combination [@RodrikD:2011].
-Using resources efficiently, fairly and with maximum freedom has been proposed as an unattainable triad [@Coyle:2011].
-Similarly, structural tradeoffs between prosperity, equality, and the environment  have been investigated in different locations using various indicators.
-A statistical analysis found that economic inequality and carbon emissions were negatively correlated in low- and middle-income countries but positively correlated in high-income countries [@Grunewald2017].
-In contrast, an earlier econometric study concluded that mitigating carbon emissions trades-off with both economic growth and social equity [@Ravallion_2000].
-And focusing on three case studies in the United States and Southeast Asia, Halpern et al.
-(2013) found that achieving biodiversity objectives is often facilitated by the inequitable distribution of resources, complicating the achievement of "triple-bottom-line solutions" in conservation planning [@Halpern_2013].
-Using an agent-based model of collective decision-making that assumes simple majority rule and takes for granted prosperity-environment conflicts, Kempf and Rossignol (2007) argued that widening inequality undermines public willingness to support environmental policies [@KEMPF_2007].
+Although global prosperity has increased in recent generations, progress has been uneven.
+Even today, a significant share of the world’s population lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]).
+Right above this derisory threshold, hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018].
+And while average incomes for certain developing countries have been converging with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened across the world [@Bourguignon:2007; @Milanovic:2016].
+This has made  inequality an urgent issue for policymakers and the general public due to  anxieties about social and political instability [@Chancel:2020; @UN2020].
+Environmental integrity forms the third, and perhaps most fundamental, dimension of the sustainability [@arrow1995economic].
+There is now a scientific consensus that human activities have had severely deleterious impacts on the biosphere, the health of which determines long-run human wellbeing [@Díaz:2019; @folke2021our].
+Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023].
 
-Despite earlier studies on potential sustainability trilemmas, few analyses have taken a dynamic global perspective.
-Many of the most critical development challenges in the Anthropocene, when human activities have become the primary drivers of planetary change, are transboundary and global in scale.
-Identifying the distribution of these challenges across countries is key to arriving at principles for collective action.
-At the same time, it is important to track how different countries have progressed or regressed with respect to key objectives over meaningful time-frames.
-In this way, country-specific as well as shared international obstacles and opportunities can be identified, helping to inform the design and implementation of policies.
-
-In this study, we ask whether there is evidence – and rationale – that the three sustainability goals of prosperity, equality, and environmental integrity can be jointly achieved by any given country using an international, time-series dataset of representative development indicators.
-In other words, this study attempts to answer the questions: To what extent does such a sustainability trilemma exist and are there inherent tradeoffs between its constituent goals?
-If such a sustainability trilemma does exist, where does it occur and why does it occur there?
+Identifying global patterns and trends in the relationships between prosperity, equality, and environmental impact can provide insights into the development challenges facing individual  countries as well as shared obstacles and opportunities internationally.
+Using a time-series dataset of representative development indicators, this study asks whether there is evidence that prosperity, equality, and environmental integrity has been jointly achieved by any country, and then explores potential mechanisms of trade-offs and synergies between the three goals.
+In other words, this study attempts to answer the questions: To what extent does a "sustainability trilemma" exist, and if does exist, where does it occur and why does it occur there?
 By providing answers to these questions, we hope to contribute to the increasingly urgent discourse on how to achieve sustainable and inclusive prosperity in an age of unprecedented planetary change.
-
-## Background
-
-For most of human history, material want, not sufficiency, was the norm for the majority of people, regardless of country [@Maddison:2007; @Harari:2014].
-Although global prosperity has increased in recent generations, progress has been uneven.
-If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium].
-And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]).
-Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and aspire for higher and more secure living standards [@TheWor:2018].
-Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an unprecedented increase of global poverty and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].
-
-At the same time, while average incomes for certain countries in what was once called the "Third World" have converged with those of advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
-
-The widespread feelings of injustice that highly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009].
-Over the past decade, inequality has once again become an increasingly urgent issue for policymakers and the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020].
-These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks [@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017].
-Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened economic inequalities within many countries, as well as widened the development trajectories between high- and low-income economies [@WorldBank:2022].
-
-Environmental integrity forms the third, and arguably most fundamental, dimension of the sustainability trilemma [@arrow1995economic].
-Interdisciplinary research confirms that human well-being and economic development depend on the resilience of nature.
-There is also a scientific consensus that human activities have a wide and deleterious impact on the biosphere, which constitutes the habitative upper bound and common denominator for our species (notwithstanding fanciful visions of extraterrestrial expansion)  [@Díaz:2019; @folke2021our].
-Anthropogenic drivers have become primarily responsible for global biodiversity loss, land degradation, and climate change, among other environmental impacts [@Díaz:2019].
-Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023].
-Global development in the 21st Century therefore urgently requires more sensible stewardship of the biosphere, from ecosystems to the climate [@Millenniu:2005; @IPCC:2021].
 
@@
 We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time.
-We further investigated whether these movements trace out typologies of country trajectories within the trilemma space.
+We further investigated whether these movements trace out typologies of country trajectories within the three-dimensional space.
 To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.
@@
 
-Our analysis of the three representative indicators and their respective thresholds found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact.
+Our analysis found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact.
 This strongly suggests that globally, there is a sustainability trilemma.
-At the country level, there were strong tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G- per-capita GNI and ecological footprint move in the same direction).
+At the country level, there were widespread tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G - per-capita GNI and ecological footprint move in the same direction).
 Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden and New Zealand) exhibited low levels of inequality in the past but then high variation and a trend towards greater inequality in more recent years.
 
-While these globally-aggregated patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies.
-To further investigate individual trajectories, we calculated the displacement within the trilemma space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and ecological footprint plane.
+While these global patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies.
+To further investigate country trajectories, we calculated the displacement within the three-dimensional space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and environmental impact plane.
 These two angles and the total distance "travelled" in the three-dimensional space help to define typologies in country trajectories.
@@
 \centering
-\includegraphics[width = 6in, height = 4in]{trilema_ensamble}
+\includegraphics[width = 7in, height = 5in]{trilema_ensamble2}
 \caption{\textbf{Countries' positions and trajectories within the trilemma space} 59 country trajectories from 1995 to 2017 are depicted.
-Colours and contours represent areas of similar values and enclose attractors within peaks.
-The insets of each graph show the direction of movement for countries during the study period.}
+Colours and contours represent areas of similar values and enclose regimes (attractors).
+The bottom panel shows the direction of movement for countries during the study period for each facet of the trilemma cube.
+Angles and distances are calculated in the three-dimensional space, but visualised in two dimensions for readability.
+The arrows shows the direction and magnitude of movement if all countries would have started at the same point of origin (to enable comparison).
+Arrow colours on the bottom panel correspond to groups found on the clustering analysis.}
 \label{fig:cube_ensamble}
@@
 
-There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
-Building on insights from feedback dynamics described in other fields of development studies and sustainability science, multiple stable states within the trilemma space are plausible if feedbacks between development drivers are present.
-Statistically, this translates into multi-modal distributions and so-called "attractors".
-The prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters.
+There are certain regions in the three-dimensional space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
+Multiple stable states can exist if feedbacks between development drivers are present.
+Statistically, this translates into multi-modal distributions and so-called "regimes".
+The prosperity-environment plane (Fig \ref{fig:cube_ensamble}A), and inequality-environment plane (Fig \ref{fig:cube_ensamble}C) have two pronounced clusters of countries (the movement of countries during the study period stayed within a relatively circumscribed space), while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has one major cluster and one looser grouping.
 Countries also showed distinctive profiles in direction of movement (Fig \ref{fig:cube_ensamble} insets).
-Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started.
+Most countries had low displacement (short arrows), which means their trajectory over the study period did not take them far from where they started.
 A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995.
 For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint worsened while per-capita GNI grew significantly.
-And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally.
-To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three trilemma indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).
+And while inequality and environmental impact fell in Ukraine, prosperity only increased marginally.
+To better understand country typologies, we clustered summary statistics (i.e., mean and standard deviations) of the three indicators, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).
 
-The trajectories of countries in the trilemma space from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A).
-First, there was a group of countries, mainly from Latin America, that were "trapped" in a region of high inequality.
-They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green).
-The second group showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow).
-The third group was composed of countries with relatively high prosperity and ecological footprint; their movements in the trilemma space were relatively less erratic, with larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).
+Combining the three planes, the trajectories of countries from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A).
+First, there was a group composed of countries with relatively high prosperity and ecological footprint; their movements were relatively less erratic and had larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).
+This cluster can be called "Regime A".
+Next, there was a group of countries, mainly from Latin America, that seemed to be "trapped" in a region of high inequality during the study period.
+They moved toward regions of higher income inequality with respect to the sample and with a larger standard deviation of $\theta$; that is to say, large variations along the income inequality and ecological footprint axes (Fig \ref{fig:pca}A, in green).
+Within the three-dimensional space, this cluster of countries with high inequality and relatively low prosperity and environmental impact can be called "Regime B".
+A third group of countries showed large displacements within the trilemma space but in no definite direction, and was characterized by large variations in prosperity and inequality (Fig \ref{fig:pca}A, in yellow).
 
-If we ignore temporal dynamics and focus on the mean and standard deviation of the trilemma indicators, we are able to include more countries (N=140).
-This static analysis yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the dynamic analysis.
+If we ignore temporal dynamics and focus on the mean and standard deviation of the indicators, we are able to include more countries (N=140).
+This static analysis also yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the findings from the dynamic analysis.
 The first cluster (Fig \ref{fig:pca}B, in yellow) was composed primarily of advanced Western economies and had high prosperity, inequality, and environmental impact.
-The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "developing world" or "Global South".
+The second cluster (Fig \ref{fig:pca}B, in green) was composed of countries from Africa, Latin America, and Asia - broadly, the "Global South".
 Here, countries all tended to have lower prosperity (e.g., low- or middle-income status), but among them exhibited large variations in inequality and environmental impact.
 
-Decreasing inequality and decreasing environmental impact tended to pull countries in opposite directions (Fig \ref{fig:pca}), providing strong evidence for the difficulty of achieving both objectives at the same time.
-In Fig \ref{fig:pca}A the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences.
-While environmental footprint (EF) and gross national income (GNI) point in similar directions, meaning countries tend to increase both at the same time, our inequality proxy (Gini) points in the opposite direction, confirming our previous speculation: countries that reduce inequality do so by increasing their environmental footprint and prosperity.
-In Fig \ref{fig:pca}B the ordination is performed in a larger but static sample (140 countries), and we find a similar pattern although not completely in opposite directions.
-In fact, in our trajectories analysis we found that  countries that reduced inequality subsequently increased their environmental impact (Fig \ref{fig:cube_ensamble}A).
-Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B).
-However, some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact - albeit at lower levels of prosperity (Fig \ref{fig:cube_ensamble}B, C).
-Overall, our findings are in line with the conventional wisdom that higher prosperity usually comes at environmental expense.
-The low frequency with which countries were able to meet even two thresholds - none in the case of the prosperity-environment relationship - is notable.
-National trajectories through the three-dimensional trilemma space also revealed that some countries struggled to escape areas characterised by high inequality, which may represent an "inequality trap" subtype of the more general sustainability trilemma.
+In both the dynamic and the static analysis, decreasing inequality and decreasing environmental impact appear to pull countries in opposite directions (Fig \ref{fig:pca}), providing evidence for the difficulty of simultaneously achieving both goals.
+In Fig \ref{fig:pca}A, the fitted vectors (arrows) show the direction in which variables explain the variance in the data and thus countries' differences.
+While per-capita ecological footprint and GNI point in similar directions (countries tended to increase both at the same time), the Gini for income points in the opposite direction.
+This suggests that countries that reduce inequality do so after increasing in environmental impact and prosperity.
+In the ordination performed in the larger static sample (Fig \ref{fig:pca}B), we find a similar pattern.
+In fact, the trajectories analysis found that countries that reduced inequality subsequently increased in environmental impact (Fig \ref{fig:cube_ensamble}A).
+Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B), although some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact (Fig \ref{fig:cube_ensamble}B, C).
 
@@
 Three clusters are identified when looking at the historical trajectory of countries (N=59) from 1995-2017 (A).
-Two clusters are identified when aggregating over time with means and standard deviations on the trilemma dimensions regardless of data completeness.
-It complements the dynamic analysis with one summary statistic over time for 140 countries (B).
+Two clusters are identified when aggregating over time with means and standard deviations regardless of data completeness.
+This complements the dynamic analysis with one summary statistic over time for 140 countries (B).
 For both A and B, red arrows are vectors fitted to the data that indicate the directions along which countries and clusters differ from each other.
-Mean values of ecological footprint (C), inequality (D), and gross national income (E) are shown in maps for the 140 countries.}
+Mean values of per-capita ecological footprint (C), Gini for income (D), and per-capita gross national income (E) are shown in maps for the 140 countries.}
 \label{fig:pca}
@@
 
-In this study, we conducted a retrospective, dynamic analysis, and it is important to note that past trajectories do not necessarily determine future pathways.
-The potential mechanisms of tradeoffs and synergies between prosperity, inequality, and environmental impact constitute an important research frontier in sustainability science and development studies.
-In the often disparate literatures of economic, social, and environmental science, no theory has so far been advanced to account for the country patterns and typologies identified here.
-Certain mechanisms have been hypothesized for pairwise relationships among the development indicators used in this study.
-However, only by considering all three dimensions together, and the different ways in which countries have evolved with respect to them, can we derive a more comprehensive picture of the sustainability challenges for international development.
-Before positing trilateral explanations for the existence and distribution of the sustainability trilemma, we will first review several of the most influential dilemma hypotheses and discuss why they may be at odds with our findings.
+Understanding the patterns and trends in the relationships between prosperity, inequality, and environmental impact is an important topic for sustainability science and development research.
+In the existing literatures of economic, social, and environmental science, various arguments have been advanced about the relationships between the three goals.
+Most commonly, mechanisms have been hypothesized for pair-wise relationships, with Kuznets Curve-type relationships being the most widely analyzed.
+The (Environmental) Kuznets Curve posits that growing prosperity leads to worsening inequality and environmental impact, before reaching an inflection point after which the relationship is reversed.
+The evidence for this hypothesis has been contested with respect to both inequality [@Chancel:2020][@Pikett:2014] and the environment [@stern2004rise; @stern1996economic ].
+For the latter, conclusions have often varied based on the indicator chosen to proxy environmental impact [@MALER201325].
+Our analysis did not detect KC-type relationships between prosperity and either the Gini coefficient for income or ecological footprint Fig \ref{fig:S5}) (please see SI for more details).
+Instead, the findings strongly suggest that growing prosperity is almost always associated with worsening inequality and environmental impact.
 
-Tradeoffs between inequality and prosperity (e.g., per-capita income) have been extensively studied in relation to the concept of the Kuznets Curve (KC).
-In this framing, the former increases as the latter grows, eventually reaching an inflection point after which growing prosperity leads to lower inequality [@Kuznets:1955].
-The evidence for this hypothesis has been contested [@Chancel:2020].
-Sustained inequality reduction over the past century seems to have been a consequence of contingent but systemic shocks - the world wars and the Great Depression - rather than through regularities of the global economy or even concerted government actions [@Pikett:2014].
-Our analysis of per-capita income and income distribution found no evidence for a KC-like relationship from 1995 to 2017, which was a period of rapid economic growth in many parts of the world (see SI).
+What is clear from earlier empirical studies of pair-wise relationships of prosperity, inequality, and the environment is that whatever tradeoffs and synergies that could be identified were mediated by exogenous factors [@dasgupta2002confronting; @pothen:2019].
+For instance, in his survey of relevant studies, Chancel [-@Chancel:2020] concludes that what empirical evidence has been found to support (E)KC-type relationships are actually artefacts particular to post-World War II datasets or transient historical trends with limited applicability to today’s inequality dynamics.
+It is only by considering the economic, social, and environmental dimensions together, and the different ways in which countries have evolved with respect to them, can we arrive at a more comprehensive understanding.
+In particular, while the prosperity-inequality and prosperity-environment relationships have been widely analyzed (with mixed conclusions), the inequality-environment dynamic has been less explored – and should also be considered with respect to the potentially mediating effects of the third dimension of prosperity.
+In our findings, this necessity is underscored by the existence of international clusters that moved in distinctively different directions.
 
-In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) is a complementary hypothesis: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence [@barbier1997economic; @grossman1995economic].
-As with the original KC, empirical studies of a wide range of variables have produced mixed findings.
-Some studies have identified evidence for this inverted-U-shaped relationship.
-For instance, EKC relationships were often found for local air pollutants but not for river pollution, deforestation, or global greenhouse gas emissions [@stern2004rise; @stern1996economic ].
-In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we did not detect an EKC-type trend: i.e., an inflection point in the prosperity-environmental impact plane (Fig \ref{fig:cube_ensamble}, Fig \ref{fig:S5}).
-Various mechanisms have been proposed to explain this relationship (or its absence), including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325].
+Overall, the low frequency with which countries were able to meet even two goals – no countries in the case of the prosperity-environment relationship – in both the dynamic and static analysis is notable.
+This suggests that there is indeed a sustainability trilemma facing global development.
+In other words, despite its prominence as a definition of sustainability, our study indicates it is highly difficult, and potentially infeasible, for countries to reach the triple bottom line of prosperity, equality, and environmental integrity.
+However, the burden of overcoming the trilemma is not even across countries.
+Another important finding of the analyses is revealed through national trajectories in the three-dimensional space.
+Over the study period, some countries, primarily from Latin America, struggled to escape areas characterised by high inequality, which may represent an "inequality trap" subtype of the trilemma.
 
-What is clear from these earlier empirical studies of pairwise relationships between prosperity, inequality, and the environment is that whatever tradeoffs and synergies that were identified are mediated by exogenous factors [@dasgupta2002confronting; @pothen:2019].
-Indeed, in his in-depth survey of relevant studies, Chancel (2020) concludes that what empirical evidence has been found to support (E)KC-type relationships are actually artefacts particular to post-World War II datasets or transient historical trends with limited applicability to today’s inequality dynamics [@Chancel:2020].
-Within the trilemma space, this need for more sophisticated explanatory frameworks was underscored by the existence of international clusters that moved in distinctively different directions.
-These dynamics highlight the inadequacy of reductionist explanations that do not incorporate institutional, historical and geographic factors and the social-ecological feedbacks that reinforce their effects.
+The existence of the trilemma in general, and of the inequality trap in particular, could be better explained through an understanding of how inequality and environmental impact are related.
+A potentially useful approach is to investigate how different levels of income distribution impact environmental integrity at different levels of average income (thereby incorporating the prosperity dimension).
+In the existing literature on inequality-environment dynamics, two broad types of mechanisms have been posited: those that are "economic" (based on assumptions about the consumption decisions of rational individuals with different relative incomes) and those that are "political" (based on assumptions about how income distribution influences collective decision-making in the form of public policy) [@berthe2015mechanisms; @@Hamann:2018].
 
-A key institutional factor driving divergent international development outcomes over our study period was the structure of global commerce.
-In particular, it has become evident that globalization created or exacerbated the (already) skewed distribution of environmental impacts worldwide.
-Notably, affluent countries "outsource" high-throughput,- land-intensive, and environmentally-polluting production to developing countries through international supply chains [@Nijdam:2005; @Peters:2006; @Lambin:2011; @Weinzett:2013] - an important factor not captured in most EKC analyses [@MALER201325].
-Additionally, it has been found that international trade influences global land use patterns, worsening natural resource exploitation and biodiversity loss [@Creutzig:2019].
-Our study period of 1995-2017 coincided with an intensive period of globalization, stimulated by the end of the Cold War and driven by the integration of large, emerging economies into international commercial and financial networks.
+The "economic" (often called the "marginal propensity to emit") hypothesis argues that higher inequality lessens environmental impact because a more equal distribution raises overall consumption levels.
+This is based on the conventional microeconomic assumption that there is a higher marginal propensity to consume among the poor than among the rich.
+The political – sometimes "political economy" – hypothesis was first posited by Boyce [-@boyce1994inequality] and argues that by concentrating power in the hands of the rich, economic inequality worsens environmental conditions.
+One mechanism is that the well-heeled tend to have a vested interest in stopping regulations and laws that protect the environment (and thereby lower profits).
+This has been supported by agent-based modelling studies such as that by Kempf and Rossignol [-@KEMPF_2007].
+Another hypothesized mechanism comes from the opposite side of the social spectrum, in which high inequality drives the poor to engage in more environmentally-destructive activities to meet their basic needs [@zhao2021analysis].
 
-With respect to historical factors, Ukraine provides a potentially instructive example.
-At the end of the study period, Ukraine had (relatively) low levels of inequality and environmental impact.
-However, the country’s  trajectory through the trilemma space was turbulent: over the study period, indicators rose and fell in dramatic downswings due to market shocks and corruption, which has been endemic in many post-Soviet countries [@Duvanova:2014], as well as the social and economic dislocations caused by violent conflict.
-It is noteworthy that it was only in 2013 that Ukraine’s ecological footprint fell below its biocapacity, accompanied, or more likely driven, by a precipitous decline in per-capita income  (Fig.
-\ref{fig:S4}).
-The importance of geographic factors is underscored by the fact that countries that performed relatively well across all three dimensions tended to be advanced Western economies, particularly those that had a much higher ratio of space and resources to people.
-This is reflected in these countries’ per-capita biocapacity, providing greater potential for the increased ecological footprint that is associated with economic growth.
+Some studies have found greater empirical evidence for the economic explanation than the political (economy) mechanisms.
+In a study of Sub-Saharan countries from 1995 to 2018, Gimba _et al._ [-@gimba2023towards] found that rising income inequality reduces ecological footprint but exerts a significant drag on economic growth – with critical implications for poverty alleviation in a region with many of the world’s lowest-income countries.
+In a study of Western African states from 1984 to 2016, Langnel _et al._ [-@langnel2021income] similarly found that income inequality often lowers ecological footprint.
+The "marginal propensity to emit" hypothesis also found empirical support in Ali [-@ali2023income], which analyzed found a tradeoff relationship between the Gini for income and carbon emissions in 42 middle-income countries between 1990 and 2016.
+In high-income economies, Hailemariam _et al._ [-@hailemariam2020carbon]’s study of OECD countries found that increasing income inequality (also measured by Gini) was associated with decreasing carbon emissions.
 
-The effects of globalization, historical contingency, and geographical constraints are perhaps best illustrated by the country that underwent the greatest magnitude of change in all three dimensions during the study period.
-In China, the rapid rise of prosperity, inequality, and environmental impact were significantly driven by the country’s integration into international circuits of trade and finance, especially after accession to the World Trade Organization in 2001 (Fig.
-\ref{fig:S4}).
-The rapid growth of incomes was underpinned by a resource-intensive export sector while rising income inequality was due in large part to geographic segmentation: urban coastal areas were the affluent export hubs while more rural inland areas supplied raw materials, intermediate goods, and cheap labour [@xie2014income; @qi2013interpreting].
-These changes, and the subsequently dramatic changes in all three sustainability dimensions, were set in motion after the tumultuous historical events of the 1960s and 1970s, which prompted the national leadership to undertake extensive economic reforms.
+However, there has also been strong empirical evidence for the opposite effect, with lower inequality corresponding to lower environmental impact.
+In a major early study, Holland _et al._ [-@holland2009cross] found that high income inequality (measured by Gini) was a significant factor predicting anthropogenic biodiversity loss.
+In a study of 135 countries over the period 1981-2012, Malerba [-@malerba2020trade] found that reducing income inequality consistently reduced the tradeoff between rising prosperity and environmental impact as measured by the "carbon intensity of poverty reduction." In a study of five middle-income ASEAN countries, Masud _et al._ [-@masud2020does] found that from 1985-2015, a tight relationship between income inequality and environmental degradation, and with a bidirectional causality for the bottom 40% of the income distribution.
+Overall, they concluded that "policymakers should reduce income inequality to strengthen environmental sustainability".
+Focusing on the relationships between carbon emissions, economic growth, and income inequality before and after the 2008 "Great Recession" across 70 countries, a "robust finding" by Agusdinata _et al._ [-@agusdinata2021economic] was that countries on trajectories of declining emissions were also on trajectories of declining inequality.
 
-Interdisciplinary studies have begun to highlight the role of social-ecological feedbacks in reinforcing the effects of historical contingency (generating hysteresis or path dependency) and geographic constraints.
-For instance, Weinzettel et al.
-[-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
-Similarly, population, environmental carrying capacity, and pollution exhibited mutually-reinforcing feedbacks across countries [@LaTorr:2019].
-This helps to explain the relatively good performance of countries such as Canada, Sweden, and Australia in our analysis.
-These types of feedbacks can create and maintain "red loop" or "green loop" dynamics between poverty and the environment, entrapping some countries in long-term conditions of high poverty and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018].
+The case for the inequality-environment relationship being segmented by development level (i.e., average national income) has also found evidence in analyses of global datasets.
+Ehigiamusoe _et al._ [-@ehigiamusoe2022nexus] found that while higher income inequality lowers carbon emissions and ecological footprints in high-income countries, it worsens them in middle-income states and the effect of poverty reduction on environmental outcomes seems to dominate in lower-income countries.
+Similarly, Grunewald [-@Grunewald2017] found that income inequality and carbon emissions were negatively correlated in low- and middle-income countries but positively correlated in high-income countries.
 
-Although this study identified important patterns of international development and posited their potential, underlying drivers, it still faced several limitations.
-First, the offshoring of environmental degradation through global supply chains discussed above is not fully captured in standard ecological footprint measurements [@VandenBergh:2012].
+Our findings tend to support the conclusion that lowering economic inequality and lowering environmental impact are synergistic.
+In particular, in our three-dimensional analyses of country trajectories within the trilemma space, no country started or ended in a region with (relatively) high inequality and high ecological footprint (the upper right quadrant of the cube).
+And based on their trajectories, all countries first decreased in income inequality before increasing in ecological footprint, regardless of prosperity level or magnitude of displacement from their starting point.
+This suggests that countries still have developmental "room" to decrease income inequality without increasing environmental impact (Figure 1G; Figure 2C).
+
+Nonetheless, we recognize that the divergent trajectories (including both direction and magnitude of displacement) of countries and pronounced clustering, as well as the large variations in indicator performance in the static analyses, highlight the importance of structural international differences.
+This applies not only to geographic factors (e.g., population density, natural resource endowments, and climatic conditions determine per-capita biocapacity) but also to past development trajectories and the position of different countries in the structure of the world economy.
+Figge _et al._ [-@figge2017effects] found that even after controlling for prosperity (average income), globalization significantly increased countries’ ecological footprints globally.
+And based on analysis of per-capita ecological footprint and Gini for income, Teixidó-Figueras and Duro [-@teixido2015building] concluded that the primary approach to reduce overall global environmental impact is to reduce the large disparities between countries’ per-capita footprints through redistributive policies across countries, prioritizing international equity over economic growth per se.
+Therefore, structural reforms to international networks of trade and finance – which has come to be increasingly advocated by scholars, activists, policymakers in recent years – could be crucial to overcoming the sustainability trilemma.
+
+Social-ecological feedbacks, which often reinforce the effects of geographic factors and development path dependencies, can also help explain the appearance of "regimes" such as the inequality trap found in the trilemma space.
+Weinzettel _et al._ [-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
+Similarly, population, environmental carrying capacity, and pollution have been found to have mutually-reinforcing feedbacks across countries [@LaTorr:2019].
+This helps to explain the relatively good performance of countries such as Canada, Sweden, and Australia in our analyses.
+These types of social-ecological feedbacks can create and maintain "red loop" or "green loop" dynamics, entrapping some countries in long-term conditions of high poverty, inequality, and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018].
+
+#Limitations and Conclusions
+
+Although this study identified new international patterns and trends in the relationships between prosperity, equality, and environmental impact, it still faced some limitations.
+First, the offshoring of environmental degradation through global supply chains is not fully captured in standard ecological footprint measurements [@VandenBergh:2012].
 A consumption-based approach to measuring environmental impacts – as opposed to the more mainstream production-based approach – may help internalize at least some of these externalities (e.g., many resource-intensive goods produced in Asia are consumed in the US and Europe, but the consequent increases in ecological footprint are attributed to the former).
@@
 Addressing this issue has important implications for the international distribution of responsibilities with respect to global commons challenges such as mitigating climate change, biodiversity loss, and pandemic risks - all of which disproportionately affect lower-income countries [@Fankhauser:2014; @Mendelsohn:2006; @IPCC:2021].
+It could also lend support for the much-needed reforms to international structures of trade and finance discussed earlier.
 
-Whether the synergies and tradeoffs between prosperity, inequality, and the environmental impact identified in this study hold across scales (e.g., at subnational scales or in other analytic units such as cities or corporations), remains an open question.
-Furthermore, the feedback mechanisms for interactions among the three dimensions at different scales are also likely to be complex rather than mono-causal [@Weinzett:2013; @Cumming:2018; @Hamann:2018; @LaTorr:2019].
-If there is a unifying principle or concern among countries and across scales, it may be the foundational role of nature, which provides the basic life-support systems for humanity and cannot be understated when thinking about sustainable development.
-As our findings strongly suggest, a myopic focus on ever-growing prosperity as the solution to inequality and degradation is misplaced - neither KC or EKC relationships were found in our analysis (SI).
-Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts.
-One undeniable trend from our analysis of recent world development is that a singular focus on economic growth can undermine progress towards social equality and, most especially, environmental integrity [@arrow2004we; @fanning2022social].
-
-Finally, we recognize that there are other valid ways to conceptualize sustainable development.
+We also recognize that there are other valid ways to conceptualize sustainable development.
 For instance, although they did not account for distributional issues, studies on the relationship between ecological limits, human population, and the economy have a long pedigree [@Ehrlich:1971; @Meadows:1972].
@@
 Such classification schemes are informative but may also be overdetermined: it is conceivable that a different and equally justifiable list of thresholds could be proposed.
-In addition, there have been calls for analyses to better integrate social and ecological goals in order to account for the foundational role played by nature [@reyers2020global].
+In addition, there have been calls for analyses to better integrate social and ecological goals to account for the foundational role played by nature [@reyers2020global].
 While acknowledging that our "classic" triple bottom line framework is imperfect, we believe it nonetheless captures long-standing and important dimensions of sustainability and how countries have progressed or regressed within the developmental space formed by them.
 
-# Conclusions
-
-Our study found that countries around the world face a trilemma when it comes to improving economic prosperity and social equality while reducing environmental impact - three goals that the global development agenda has identified as being fundamental to sustainability.
-It is important to recognize how globalization, historical contingencies, and geographic constraints d - along with the feedbacks that strengthen their effects -  vary across countries and drive their development within the trilemma space.
-When it comes to such complex development challenges in an era of planetary change, there are no "one-size-fits-all" solutions [@Ostrom:2007; @ostrom2010moving; @perrings2007future].
-We therefore argue that national approaches to overcoming the sustainability trilemma will need to be context-specific and must account for path-dependence.
-At the same time, our findings indicate that there are typologies of how countries have historically evolved with respect to prosperity, inequality, and the environment.
-These typologies can help identify common priorities and suggest development strategies.
-This may be particularly relevant for Global South countries, which appear to face the steepest challenges with respect to the sustainability trilemma.
-
-Overcoming the sustainability trilemma, for individual countries and the world as a whole, demands development that focuses as much on the distribution and impact of the economy as on the economy’s continued growth.
-This will likely require much greater investment in the social and natural capital that have been degraded, or allowed to atrophy, across the world in recent decades, from advanced Western economies to the large "emerging markets" to the least developed states of the Global South - where healthy communities and resilient environments will be critical for their future in the Anthropocene [@pretty2003social; @brondizio2009connectivity; @wackernagel1997perceptual; @Ehrlich:2012].
-Through this more inclusive approach, countries could still become more prosperous.
-But that prosperity should be achieved in ways that distributes gains more fairly, which would help resolve the poverty and precarity that still bedevil so many, and with less impact on the biosphere, which remains the ultimate foundation for humanity’s prospective flourishing.
+Whether the synergies and tradeoffs between prosperity, inequality, and the environmental impact identified in this study hold across scales (e.g., at subnational scales or in other analytic units such as cities or corporations), also remains an open question.
+Furthermore, the feedback mechanisms for interactions among the three dimensions at different scales are also likely to be complex [@Weinzett:2013; @Cumming:2018; @Hamann:2018; @LaTorr:2019].
+If there is a unifying concern among countries and across scales, it may be the foundational role of nature, which provides the basic life-support systems for human well-being and the inescapable context for collective development.
+As our findings strongly suggest, a myopic focus on ever-growing prosperity as the solution to inequality and degradation is misplaced.
+Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts.
+One undeniable trend identified in our analysis is that a singular focus on economic growth can undermine progress towards social equality and, most especially, environmental integrity [@arrow2004we; @fanning2022social].
 

### 2023-11-20 figures.R

@@ -1,0 +1,635 @@
+library(tidyverse)
+library(patchwork)
+library(plot3D)
+library(magrittr)
+
+
+load("data/cleaned_trilemma-20230614.RData")
+load("data/trilemma_static.Rda")
+
+#### Conceptual figure ####
+old_par <- par()
+
+quartz(width = 2.5, height = 2.5, pointsize = 7)
+par(mar = c(0,0,0,0))
+box3D(
+    x0 = -0.8, y0 = -0.8, z0 = -0.8, 
+    x1 = 0.8, y1 = 0.8, z1 = 0.8, 
+    border = "black", lwd = 1, 
+    col = gg.col(1, alpha = 0.25), 
+    xlab = "Inequality", ylab = "Environment",
+    zlab = "Prosperity", cex.lab = 1.2, mar = c(0,0,0,0))
+
+quartz.save(
+    file="figures/cube_concept.png", type = "png", dpi = 300,
+    width = 2.5, height = 2.5, pointsize = 7, bg = "white"
+)
+dev.off()
+par(old_par)
+
+
+p1 <- ggplot() +
+    geom_blank() + 
+    geom_rect(aes(xmin = 0, xmax = 0.5, ymin = 0, ymax = 0.5), 
+              fill = "skyblue", alpha = 0.5) +
+    geom_rect(aes(xmin = 0, xmax = 0.5, ymin = 0.5, ymax = 1), 
+              fill = "orange", alpha = 0.5) +
+    geom_rect(aes(xmin = 0.5, xmax = 1, ymin = 0, ymax = 0.5), 
+              fill = "orange", alpha = 0.5) +
+    geom_rect(aes(xmin = 0.5, xmax = 1, ymin = 0.5, ymax = 1), 
+              fill = "red", alpha = 0.5) +
+    geom_abline(intercept = 0, slope = 1, color = "blue") +
+    geom_abline(intercept = 1, slope = -1, color = "red") +
+    geom_hline(yintercept = 0.5, linetype = 2, color = "grey50", size = 0.5) +
+    geom_vline(xintercept = 0.5, linetype = 2, color = "grey50", size = 0.5) +
+    annotate("text", x = 0.15, y = 0.2, label = "Synergies", size = 2,
+             color = "blue", angle = 45) +
+    annotate("text", x = 0.15, y = 0.9, label = "Trade-off", size = 2,
+             color = "red", angle = 360-45) +
+    lims(x = c(0,1), y = c(0,1)) +
[…]

### 2023-11-20 submit.R

@@ -10,3 +10,8 @@ pdf_subset("paper/manuscript.pdf", pages = 1, output = "paper/conflict_statement
 pdf_subset("revised_manuscript_230703.pdf", pages = 4:32, output = "revised_main_manuscript.pdf")
 pdf_subset("response to reviewers.pdf", pages = 1, output = "letter_editor.pdf")
 pdf_subset("response to reviewers.pdf", pages = 2:7, output = "response_reviewers.pdf" )
+
+
+## Third submission
+pdf_subset("response_letter-reviewers.pdf", pages = 1, output = "revision_231118/response_letter.pdf")
+pdf_subset("response_letter-reviewers.pdf", pages = 2:4, output = "revision_231118/response_reviewer.pdf")

### 2024-02-26 revised_manuscript_240226.Rmd
revisions round 4

--- 2023-11-20 revised_manuscript_231115.Rmd
@@
 - No country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact between 1995-2017.
-- A sustainability “trilemma” as opposed a “triple bottom line” appears to exist for international development.
-- There is an “inequality trap” that appears to be particularly acute in Latin America.
-- Country trajectories suggest that reducing inequality can be achieved without compromising the biosphere.
-- The three dimensions need be considered together to identify international development challenges.
+- There appears to be a global sustainability "trilemma" as opposed to a "triple bottom line".
+- This global challenge was unevenly distributed, evidenced by countries’ divergent trajectories and clustering effects.
+- There were widespread tradeoffs among the three indicators, with the prosperity-environment tradeoff most pronounced.
+- Mitigating the prosperity-environment tradeoff is made more difficult by high inequality.
 
@@
 A key aim of sustainable development is the joint achievement of prosperity, equality, and environmental integrity: in other words, material living standards that are high, broadly-distributed, and low-impact.
-This has often been called the “triple bottom line”.
-But instead, what if there is a " trilemma" that inhibits the simultaneous achievement of these three goals?
-We analyzed international patterns and trends in the relationships between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017.
-Based on these widely-used measures of development, we found that no country achieved the triple bottom line.
-Over the study period, countries tended to decrease in income inequality before growing in prosperity and environmental impact.
-A cluster of countries, primarily from Latin America, appeared trapped in a regime of high inequality, while the instances of prosperous countries lowering inequality only coincided with increasing environmental impact.
-The existence of country clusters and widely-divergent national trajectories show that common pair-wise explanations such as Kuznets Curves do not fully capture important development dynamics, especially in a period of global environmental change.
-Instead, frameworks need to consider all three dimensions together, and especially the relationship between inequality and the environment.
-Additionally, the past development trajectories of countries, how they fit into the structure of the world economy, and the role of social-ecological feedbacks in strengthening clustering effects such as “inequality traps”.
-Practically, our findings suggest that to overcome the sustainability trilemma, many countries - especially those that are already prosperous - should prioritise social redistribution and environmental stewardship before further economic growth.
+This has often been called the "triple bottom line".
+But instead, what if there is a "trilemma" that inhibits the simultaneous achievement of these three goals?
+We analyzed international patterns and trends in the relationships between per-capita gross national income, the Gini coefficient for income distribution, and per-capita ecological footprint from 1995 to 2017, benchmarking them against thresholds from the existing literature.
+A "dynamic" analysis of the trajectories of 59 countries and a "static" analysis of a larger sample of 140 countries found that none met the triple bottom line, and that instead there were widespread tradeoffs among the three indicators.
+These tradeoffs, leading to divergent national trajectories and country clusters, show that common pair-wise explanations such as Kuznets Curves do not adequately capture important development dynamics.
+In particular, while only a few countries simultaneously met the thresholds for prosperity and equality on the one hand and equality and environment on the other, none did for prosperity and environment.
+Moreover, inequality likely makes resolving this critical tradeoff more difficult.
+Our findings suggest that mitigating the sustainability trilemma may require countries - especially those that are already prosperous - to prioritise economic redistribution and environmental stewardship over further growth.
 
@@
 Environmental integrity forms the third, and perhaps most fundamental, dimension of the sustainability [@arrow1995economic].
-There is now a scientific consensus that human activities have had severely deleterious impacts on the biosphere, the health of which determines long-run human wellbeing [@Díaz:2019; @folke2021our].
+There is now a scientific consensus that human activities have had severely deleterious impacts on the biosphere, the health of which determines long-run human wellbeing [@Diaz:2019; @folke2021our].
 Indeed, humanity has already exceeded a number of the biophysical boundaries that delimit a "safe operating space" for humanity [@Rockström:2009; @RaworthKate:2017; @Rockstr_m_2023].
@@
 We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time.
-We further investigated whether these movements trace out typologies of country trajectories within the three-dimensional space.
+We further investigated whether these movements trace out typologies of country trajectories within the "trilemma space".
 To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.
@@
 
-While we were able to assess 140 countries "statically" (i.e., ignoring temporal dynamics), an analysis of the missing values showed that analyzing changes over time in the trilemma space was only possible for 59 countries.
+While we were able to assess 140 countries "statically" (i.e., ignoring temporal dynamics), an analysis of the missing values showed that a "dynamic" analysis of changes over time in the trilemma space was only possible for only 59 countries (i.e., data for all three indicators were available for every year of the study period).
 The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values.
@@
 
+Finally, while global patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies.
+To further investigate country trajectories, we also calculated their displacement within the trilemma space for each pair-wise relationship, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and environmental impact plane.
+These two angles and the total distance "travelled" in the three-dimensional space help to define the typologies of country trajectories.
+
 # Results
 
-Our analysis found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact.
-This strongly suggests that globally, there is a sustainability trilemma.
-At the country level, there were widespread tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreases, ecological footprint increases) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G - per-capita GNI and ecological footprint move in the same direction).
-Results show synergies between increasing prosperity and decreasing inequality (Fig \ref{fig:trilemma}G), although a number of high-income countries (e.g., Sweden and New Zealand) exhibited low levels of inequality in the past but then high variation and a trend towards greater inequality in more recent years.
-
-While these global patterns provide insight into international variations in key dimensions of sustainability, they do not completely reveal the diversity of individual country trajectories, nor their typologies.
-To further investigate country trajectories, we calculated the displacement within the three-dimensional space for each pair-wise dimension, where $\gamma$ (gamma) is the angle in the inequality and prosperity plane while $\theta$ (theta) is the angle in the inequality and environmental impact plane.
-These two angles and the total distance "travelled" in the three-dimensional space help to define typologies in country trajectories.
+Our analyses found that no country simultaneously achieved high prosperity, low inequality, and low environmental impact.
+This strongly suggests that there is a sustainability trilemma at the global scale.
+More specifically, the dynamic analysis of 59 countries found that there were widespread tradeoffs between reducing inequality and reducing environmental impact (Fig \ref{fig:trilemma}E - as the Gini coefficient for income decreased, ecological footprint increased) as well as between increasing prosperity and reducing environmental impact (Fig \ref{fig:trilemma}G - per-capita GNI and ecological footprint moved in the same direction).
+The results also found persistent tradeoffs between prosperity and inequality (Fig \ref{fig:trilemma}G).
+Even several high-income countries (e.g., Sweden and New Zealand) that exhibited low levels of inequality in the past began to exhibit high variation and then a trend towards greater inequality with continued income growth.
 
@@
 
-There are certain regions in the three-dimensional space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
+There are certain regions in the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}).
 Multiple stable states can exist if feedbacks between development drivers are present.
@@
 A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995.
-For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint worsened while per-capita GNI grew significantly.
+For example, between 1995 and 2017, China's Gini coefficient for income and per-capita ecological footprint rose significantly alongside per-capita GNI.
 And while inequality and environmental impact fell in Ukraine, prosperity only increased marginally.
@@
 
-Combining the three planes, the trajectories of countries from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A).
+The trajectories in the trilemma space of the 59 countries from 1995 to 2017 can be categorized into three general types (Fig \ref{fig:pca}A).
 First, there was a group composed of countries with relatively high prosperity and ecological footprint; their movements were relatively less erratic and had larger values but lower variation in direction ($\gamma$ and $\theta$ angles, Fig \ref{fig:pca}A, blue group).
@@
 
-If we ignore temporal dynamics and focus on the mean and standard deviation of the indicators, we are able to include more countries (N=140).
+If we ignore temporal dynamics and focus on the mean and standard deviation of the indicators over the study period, we are able to include more countries (N=140).
 This static analysis also yielded two distinct clusters of countries (Fig \ref{fig:pca}B), with similar tradeoffs between prosperity, inequality, and environmental impact as in the findings from the dynamic analysis.
@@
 While per-capita ecological footprint and GNI point in similar directions (countries tended to increase both at the same time), the Gini for income points in the opposite direction.
-This suggests that countries that reduce inequality do so after increasing in environmental impact and prosperity.
+This shows that a few  countries that reduce inequality did so only after increasing in environmental impact and prosperity.
 In the ordination performed in the larger static sample (Fig \ref{fig:pca}B), we find a similar pattern.
 In fact, the trajectories analysis found that countries that reduced inequality subsequently increased in environmental impact (Fig \ref{fig:cube_ensamble}A).
-Reducing inequality was often followed by rising prosperity (Fig \ref{fig:cube_ensamble}B), although some middle-income countries maintained relatively low levels of inequality (at levels similar to higher-income countries) while still maintaining lower levels of environmental impact (Fig \ref{fig:cube_ensamble}B, C).
+Overall, the findings of both the dynamic and the static analysis show that there are tradeoffs between the three "bottom lines" of prosperity, equality, and environmental integrity at the global level.
+This is perhaps most simply illustrated by the fact that in both analyses, only a small subset of countries simultaneously met the prosperity and inequality thresholds on the one hand and the inequality and environment thresholds on the other, and no countries simultaneously met the prosperity and environment thresholds (Fig \ref{fig:trilemma} E-G; Fig \ref{fig:cube_ensamble} A-C).
 
@@
 
-Understanding the patterns and trends in the relationships between prosperity, inequality, and environmental impact is an important topic for sustainability science and development research.
-In the existing literatures of economic, social, and environmental science, various arguments have been advanced about the relationships between the three goals.
+Various arguments have been advanced to explain the relationships between prosperity, equality, and environmental integrity in global development.
 Most commonly, mechanisms have been hypothesized for pair-wise relationships, with Kuznets Curve-type relationships being the most widely analyzed.
-The (Environmental) Kuznets Curve posits that growing prosperity leads to worsening inequality and environmental impact, before reaching an inflection point after which the relationship is reversed.
-The evidence for this hypothesis has been contested with respect to both inequality [@Chancel:2020][@Pikett:2014] and the environment [@stern2004rise; @stern1996economic ].
-For the latter, conclusions have often varied based on the indicator chosen to proxy environmental impact [@MALER201325].
-Our analysis did not detect KC-type relationships between prosperity and either the Gini coefficient for income or ecological footprint Fig \ref{fig:S5}) (please see SI for more details).
+The (Environmental) Kuznets Curve posits that growing prosperity leads to worsening inequality and environmental impact, before reaching an inflection point after which the is reversed.
+The evidence for this hypothesis has been contested with respect to both inequality [@Chancel:2020; @Pikett:2014] and the environment [@stern2004rise; @stern1996economic ].
+In a survey of the empirical literature, Chancel [-@Chancel:2020] concluded that evidence supporting (Environmental) Kuznets Curve-type relationships tended to be statistical artefacts particular to post-Second World War datasets or transient historical trends.
+Therefore, such findings have limited applicability to today’s conditions of large-scale social-ecological change.
+
+Our findings support these arguments: neither the dynamic nor the static analysis detected Kuznets Curve relationships between prosperity and either the Gini coefficient for income or ecological footprint Fig \ref{fig:S5}) (please see SI for more details).
 Instead, the findings strongly suggest that growing prosperity is almost always associated with worsening inequality and environmental impact.
+The low frequency with which countries were able to meet even two goals in both the dynamic and the static analysis is highly notable.
+Indeed, the fact that no country simultaneously met the prosperity and environment thresholds suggests that resolving the "dilemma" of this particular tradeoff is the necessary (but insufficient) condition for resolving the broader sustainability trilemma facing global development.
 
-What is clear from earlier empirical studies of pair-wise relationships of prosperity, inequality, and the environment is that whatever tradeoffs and synergies that could be identified were mediated by exogenous factors [@dasgupta2002confronting; @pothen:2019].
-For instance, in his survey of relevant studies, Chancel [-@Chancel:2020] concludes that what empirical evidence has been found to support (E)KC-type relationships are actually artefacts particular to post-World War II datasets or transient historical trends with limited applicability to today’s inequality dynamics.
-It is only by considering the economic, social, and environmental dimensions together, and the different ways in which countries have evolved with respect to them, can we arrive at a more comprehensive understanding.
-In particular, while the prosperity-inequality and prosperity-environment relationships have been widely analyzed (with mixed conclusions), the inequality-environment dynamic has been less explored – and should also be considered with respect to the potentially mediating effects of the third dimension of prosperity.
-In our findings, this necessity is underscored by the existence of international clusters that moved in distinctively different directions.
+A growing body of empirical studies have sought to identify potential "decoupling" between economic growth and various indicators of environmental impact.
+The evidence for such trends have been mixed.
+In a review of the literature going back thirty years, Vadén _et al_.
+[-@vaden2020decoupling] found evidence of global decoupling between economic growth and certain pollutants, as well as country-level cases of decoupling from land and water use.
+But evidence of economy-wide decoupling from resource use at the international scale has proven elusive.
+However, Szigeti et al.
+[-@szigeti2017decoupling] did identify a positive international trend in the decreasing ecological footprint intensity of economic growth over the past two decades.
+But whether this trend can continue to the point where the prosperity-environment tradeoff is fully mitigated remains an open and pressing question.
+Our findings suggest that this dilemma could remain a major hurdle for global development into the foreseeable future.
+Breaking the link will likely require large-scale changes in governance and cultural norms, including replacing the policy primacy of economic growth with the prioritization of social and environmental goals, especially in already affluent communities and countries [@wiedmann2020scientists].
 
-Overall, the low frequency with which countries were able to meet even two goals – no countries in the case of the prosperity-environment relationship – in both the dynamic and static analysis is notable.
-This suggests that there is indeed a sustainability trilemma facing global development.
-In other words, despite its prominence as a definition of sustainability, our study indicates it is highly difficult, and potentially infeasible, for countries to reach the triple bottom line of prosperity, equality, and environmental integrity.
-However, the burden of overcoming the trilemma is not even across countries.
-Another important finding of the analyses is revealed through national trajectories in the three-dimensional space.
-Over the study period, some countries, primarily from Latin America, struggled to escape areas characterised by high inequality, which may represent an "inequality trap" subtype of the trilemma.
+Our findings also show that inequality is likely to make mitigating prosperity-environment tradeoffs more difficult, as it appears to "pull" country trajectories away from the prosperity and environment thresholds in the trilemma space (Fig \ref{fig:trilemma}E-G).
+With respect to the prosperity dimension, there is strong empirical evidence showing that economic inequality tends to slow growth for a variety of institutional and behavioural reasons, from rent-seeking and the political influence of vested interests to the basic microeconomic logic that the marginal propensity to consume is inversely related to affluence.
+In other words, a more unequal economy is likely to have relatively lower aggregate demand [@stiglitz2016inequality; @carvalho2016personal; @aguiar2015has].
+The widespread tradeoffs between per-capita GNI and the Gini coefficient for income discovered in our analyses support this conclusion.
 
-The existence of the trilemma in general, and of the inequality trap in particular, could be better explained through an understanding of how inequality and environmental impact are related.
-A potentially useful approach is to investigate how different levels of income distribution impact environmental integrity at different levels of average income (thereby incorporating the prosperity dimension).
-In the existing literature on inequality-environment dynamics, two broad types of mechanisms have been posited: those that are "economic" (based on assumptions about the consumption decisions of rational individuals with different relative incomes) and those that are "political" (based on assumptions about how income distribution influences collective decision-making in the form of public policy) [@berthe2015mechanisms; @@Hamann:2018].
+The relationship between inequality and the environment is more complicated.
+Two primary types of mechanisms have been posited in the literature, one "economic" and the other "political" [@berthe2015mechanisms; @Hamann:2018].
+Similar to inequality’s effect on economic growth, the former hypothesizes that higher inequality lessens environmental impact because it reduces overall consumption (this has sometimes been called the "marginal propensity to emit" effect).
+The political hypothesis argues that, by concentrating power in the hands of the rich, inequality worsens environmental conditions because the well-heeled tend to have a vested interest in stopping profit-lowering regulations that protect the environment [@boyce1994inequality; @KEMPF_2007].
+The tradeoffs between income inequality and environmental impact in our findings support the economic hypothesis.
 
-The "economic" (often called the "marginal propensity to emit") hypothesis argues that higher inequality lessens environmental impact because a more equal distribution raises overall consumption levels.
-This is based on the conventional microeconomic assumption that there is a higher marginal propensity to consume among the poor than among the rich.
-The political – sometimes "political economy" – hypothesis was first posited by Boyce [-@boyce1994inequality] and argues that by concentrating power in the hands of the rich, economic inequality worsens environmental conditions.
-One mechanism is that the well-heeled tend to have a vested interest in stopping regulations and laws that protect the environment (and thereby lower profits).
-This has been supported by agent-based modelling studies such as that by Kempf and Rossignol [-@KEMPF_2007].
-Another hypothesized mechanism comes from the opposite side of the social spectrum, in which high inequality drives the poor to engage in more environmentally-destructive activities to meet their basic needs [@zhao2021analysis].
+In the existing literature, these tradeoffs have been shown for countries across the geographic and development spectrum.
+In a study of Western African states from 1984 to 2016, Langnel _et al._ [-@langnel2021income] found that income inequality often lowers ecological footprint, and Ali [-@ali2023income] discovered a tradeoff between the Gini for income and carbon emissions in 42 middle-income countries between 1990 and 2016.
+In high-income economies, Hailemariam _et al._ [-@hailemariam2020carbon]’s study of OECD countries found that increasing income inequality (also measured by the Gini coefficient) was associated with decreasing carbon emissions.
+Similar to our findings, but only focused on a subset of Sub-Saharan countries from 1995 to 2018, Gimba _et al._ [-@gimba2023towards] found that rising income inequality reduces ecological footprint but exerts a significant drag on economic growth – with critical implications for poverty alleviation in a region with many of the world’s lowest-income countries.
 
-Some studies have found greater empirical evidence for the economic explanation than the political (economy) mechanisms.
-In a study of Sub-Saharan countries from 1995 to 2018, Gimba _et al._ [-@gimba2023towards] found that rising income inequality reduces ecological footprint but exerts a significant drag on economic growth – with critical implications for poverty alleviation in a region with many of the world’s lowest-income countries.
-In a study of Western African states from 1984 to 2016, Langnel _et al._ [-@langnel2021income] similarly found that income inequality often lowers ecological footprint.
-The "marginal propensity to emit" hypothesis also found empirical support in Ali [-@ali2023income], which analyzed found a tradeoff relationship between the Gini for income and carbon emissions in 42 middle-income countries between 1990 and 2016.
-In high-income economies, Hailemariam _et al._ [-@hailemariam2020carbon]’s study of OECD countries found that increasing income inequality (also measured by Gini) was associated with decreasing carbon emissions.
+Our analyses showed that the development burden of the sustainability trilemma was distributed highly unevenly across countries.
+For instance, although no country simultaneously met the triple bottom line thresholds, some improved along two dimensions and a few – notably Belgium and New Zealand – advanced in all three.
+This heterogeneity of international performance is also underscored by the existence of country clusters or regimes (Fig \ref{fig:cube_ensamble}).
+Most notably, a group of countries primarily from Latin America struggled to escape an area in the trilemma space characterised by high inequality ("Regime 2"); this seems to have also hindered their progress in the other two dimensions, further supporting the idea that inequality makes mitigating the prosperity-environment tradeoff more difficult.
 
-However, there has also been strong empirical evidence for the opposite effect, with lower inequality corresponding to lower environmental impact.
-In a major early study, Holland _et al._ [-@holland2009cross] found that high income inequality (measured by Gini) was a significant factor predicting anthropogenic biodiversity loss.
-In a study of 135 countries over the period 1981-2012, Malerba [-@malerba2020trade] found that reducing income inequality consistently reduced the tradeoff between rising prosperity and environmental impact as measured by the "carbon intensity of poverty reduction." In a study of five middle-income ASEAN countries, Masud _et al._ [-@masud2020does] found that from 1985-2015, a tight relationship between income inequality and environmental degradation, and with a bidirectional causality for the bottom 40% of the income distribution.
-Overall, they concluded that "policymakers should reduce income inequality to strengthen environmental sustainability".
-Focusing on the relationships between carbon emissions, economic growth, and income inequality before and after the 2008 "Great Recession" across 70 countries, a "robust finding" by Agusdinata _et al._ [-@agusdinata2021economic] was that countries on trajectories of declining emissions were also on trajectories of declining inequality.
+Two plausible sets of factors may explain this unevenly-distributed burden of the sustainability trilemma: geographic heterogeneity and past development trajectories.
+Weinzettel_et al.
+[-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving ecological conditions.
+Similarly, population, environmental carrying capacity, and pollution have been found to exert mutually-reinforcing feedbacks internationally [@LaTorr:2019].
+These types of social-ecological feedbacks can create "red loop" or "green loop" dynamics, entrapping some countries in conditions of high poverty, inequality, and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018].
+In our analyses, these mechanisms may help explain the relatively good performance of certain affluent economies such as Canada, Sweden, and Australia on the one hand and the inequality-induced clustering of Latin American countries on the other.
 
-The case for the inequality-environment relationship being segmented by development level (i.e., average national income) has also found evidence in analyses of global datasets.
-Ehigiamusoe _et al._ [-@ehigiamusoe2022nexus] found that while higher income inequality lowers carbon emissions and ecological footprints in high-income countries, it worsens them in middle-income states and the effect of poverty reduction on environmental outcomes seems to dominate in lower-income countries.
-Similarly, Grunewald [-@Grunewald2017] found that income inequality and carbon emissions were negatively correlated in low- and middle-income countries but positively correlated in high-income countries.
-
-Our findings tend to support the conclusion that lowering economic inequality and lowering environmental impact are synergistic.
-In particular, in our three-dimensional analyses of country trajectories within the trilemma space, no country started or ended in a region with (relatively) high inequality and high ecological footprint (the upper right quadrant of the cube).
-And based on their trajectories, all countries first decreased in income inequality before increasing in ecological footprint, regardless of prosperity level or magnitude of displacement from their starting point.
-This suggests that countries still have developmental "room" to decrease income inequality without increasing environmental impact (Figure 1G; Figure 2C).
-
-Nonetheless, we recognize that the divergent trajectories (including both direction and magnitude of displacement) of countries and pronounced clustering, as well as the large variations in indicator performance in the static analyses, highlight the importance of structural international differences.
-This applies not only to geographic factors (e.g., population density, natural resource endowments, and climatic conditions determine per-capita biocapacity) but also to past development trajectories and the position of different countries in the structure of the world economy.
-Figge _et al._ [-@figge2017effects] found that even after controlling for prosperity (average income), globalization significantly increased countries’ ecological footprints globally.
-And based on analysis of per-capita ecological footprint and Gini for income, Teixidó-Figueras and Duro [-@teixido2015building] concluded that the primary approach to reduce overall global environmental impact is to reduce the large disparities between countries’ per-capita footprints through redistributive policies across countries, prioritizing international equity over economic growth per se.
-Therefore, structural reforms to international networks of trade and finance – which has come to be increasingly advocated by scholars, activists, policymakers in recent years – could be crucial to overcoming the sustainability trilemma.
-
-Social-ecological feedbacks, which often reinforce the effects of geographic factors and development path dependencies, can also help explain the appearance of "regimes" such as the inequality trap found in the trilemma space.
-Weinzettel _et al._ [-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions.
-Similarly, population, environmental carrying capacity, and pollution have been found to have mutually-reinforcing feedbacks across countries [@LaTorr:2019].
-This helps to explain the relatively good performance of countries such as Canada, Sweden, and Australia in our analyses.
-These types of social-ecological feedbacks can create and maintain "red loop" or "green loop" dynamics, entrapping some countries in long-term conditions of high poverty, inequality, and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018].
+Finally, the relative positions of countries in the structure of the world economy can also shed light on divergent international trajectories with respect to the three indicators.
+In an earlier study, Figge _et al._ [-@figge2017effects] found that even after controlling for average income, globalization significantly increased countries’ ecological footprints.
+And based on analysis of per-capita ecological footprint and the Gini coefficient for income, Teixidó-Figueras and Duro [-@teixido2015building] concluded that reducing global environmental impact requires reducing the large disparities between countries, including through redistributive policies that prioritize international equity over the growth of individual economies.
+Therefore, in addition to domestic policies that place greater emphasis on equality and environmental integrity, structural reforms to international networks of trade and finance may also be needed to mitigate the sustainability trilemma.
 
@@
 
-Although this study identified new international patterns and trends in the relationships between prosperity, equality, and environmental impact, it still faced some limitations.
+This study faced several limitations.
 First, the offshoring of environmental degradation through global supply chains is not fully captured in standard ecological footprint measurements [@VandenBergh:2012].
@@
 While acknowledging that our "classic" triple bottom line framework is imperfect, we believe it nonetheless captures long-standing and important dimensions of sustainability and how countries have progressed or regressed within the developmental space formed by them.
-
-Whether the synergies and tradeoffs between prosperity, inequality, and the environmental impact identified in this study hold across scales (e.g., at subnational scales or in other analytic units such as cities or corporations), also remains an open question.
-Furthermore, the feedback mechanisms for interactions among the three dimensions at different scales are also likely to be complex [@Weinzett:2013; @Cumming:2018; @Hamann:2018; @LaTorr:2019].
-If there is a unifying concern among countries and across scales, it may be the foundational role of nature, which provides the basic life-support systems for human well-being and the inescapable context for collective development.
-As our findings strongly suggest, a myopic focus on ever-growing prosperity as the solution to inequality and degradation is misplaced.
-Indeed, highly prosperous countries do not necessarily have lower levels of inequality, but almost always have higher environmental impacts.
-One undeniable trend identified in our analysis is that a singular focus on economic growth can undermine progress towards social equality and, most especially, environmental integrity [@arrow2004we; @fanning2022social].
 
@@
 \centering
-\includegraphics[width = 6in, height = 5in]{FigS4}
-\caption{\textbf{Trajectories examples} As shown by the experiences of Ukraine and China from 1995 to 2017, historical and structural factors influence the trajectories of countries within the "trilemma space" formed by the economic, social, and environmental dimensions of sustainability: (A) the global financial crisis; (B) the beginning of the Russo-Ukrainian conflict; (C) the accession of China into the World Trade Organization; (D) the beginning of the large-scale anti-corruption crackdown in China.}
-\label{fig:S4}
-\end{figure*}
-```
-
-```{=tex}
-\begin{figure*}[ht]
-\centering
 \includegraphics[width = 6in, height = 3.5in]{sm_trajectories}

### 2024-02-26 submit.R

@@ -15,3 +15,8 @@ pdf_subset("response to reviewers.pdf", pages = 2:7, output = "response_reviewer
 ## Third submission
 pdf_subset("response_letter-reviewers.pdf", pages = 1, output = "revision_231118/response_letter.pdf")
 pdf_subset("response_letter-reviewers.pdf", pages = 2:4, output = "revision_231118/response_reviewer.pdf")
+
+
+## Fourth submission
+pdf_subset("Response_Feb2024version.pdf", pages = 1, output = "revision_240225/response_letter.pdf")
+pdf_subset("Response_Feb2024version.pdf", pages = 2:4, output = "revision_240225/response_reviewer.pdf")
